\documentclass[a4paper]{book}
\usepackage{hyperref}
\usepackage{makeidx}
\usepackage{amssymb}
\usepackage{color}
\usepackage{alltt}
\usepackage{graphicx}
\usepackage{layout}
\def\union{\cup}
\def\intersect{\cap}
\def\getsrandom{\stackrel{\rm R}{\gets}}
\def\cross{\times}
\def\cat{\hspace{0.5em} \| \hspace{0.5em}}
\def\catn{$\|$}
\def\divides{\hspace{0.3em} | \hspace{0.3em}}
\def\nequiv{\not\equiv}
\def\approx{\raisebox{0.2ex}{\mbox{\small $\sim$}}}
\def\lcm{{\rm lcm}}
\def\gcd{{\rm gcd}}
\def\log{{\rm log}}
\def\ord{{\rm ord}}
\def\abs{{\mathit abs}}
\def\rep{{\mathit rep}}
\def\mod{{\mathit\ mod\ }}
\renewcommand{\pmod}[1]{\ ({\rm mod\ }{#1})}
\newcommand{\floor}[1]{\left\lfloor{#1}\right\rfloor}
\newcommand{\ceil}[1]{\left\lceil{#1}\right\rceil}
\def\Or{{\rm\ or\ }}
\def\And{{\rm\ and\ }}
\def\iff{\hspace{1em}\Longleftrightarrow\hspace{1em}}
\def\implies{\Rightarrow}
\def\undefined{{\rm ``undefined"}}
\def\Proof{\vspace{1ex}\noindent {\bf Proof:}\hspace{1em}}
\let\oldphi\phi
\def\phi{\varphi}
\def\Pr{{\rm Pr}}
\newcommand{\str}[1]{{\mathbf{#1}}}
\def\F{{\mathbb F}}
\def\N{{\mathbb N}}
\def\Z{{\mathbb Z}}
\def\R{{\mathbb R}}
\def\C{{\mathbb C}}
\def\Q{{\mathbb Q}}

\def\twiddle{\raisebox{0.3ex}{\mbox{\tiny $\sim$}}}

\def\gap{\vspace{0.5ex}}
\makeindex
\begin{document}
\title{LibTomCrypt \\ Version 0.99}
\author{Tom St Denis \\
\\
tomstdenis@iahu.ca \\
http://libtomcrypt.org
}
\maketitle
This text and source code library are both hereby placed in the public domain.  This book has been 
formatted for A4 paper using the \LaTeX{} {\em book} macro package.

\vspace{10cm}

\begin{flushright}Open Source.  Open Academia.  Open Minds.

\mbox{ }

Tom St Denis,

Phone: 1-613-836-3160

111 Banning Rd 

Kanata, Ontario 

K2L 1C3 

Canada
\end{flushright}
\newpage
\tableofcontents
\chapter{Introduction}
\section{What is the LibTomCrypt?}
LibTomCrypt is a portable ANSI C cryptographic library that supports symmetric ciphers, one-way hashes, 
pseudo-random number generators, public key cryptography (via RSA,DH or ECC/DH) and a plethora of support 
routines.  It is designed to compile out of the box with the GNU C Compiler (GCC) version 2.95.3 (and higher) 
and with MSVC version 6 in win32.

The library has been successfully tested on quite a few other platforms ranging from the ARM7TDMI in a 
Gameboy Advanced to various PowerPC processors and even the MIPS processor in the PlayStation 2.  Suffice it
to say the code is portable.

The library is designed so new ciphers/hashes/PRNGs can be added at runtime and the existing API (and helper API functions) will 
be able to use the new designs automatically.  There exist self-check functions for each cipher and hash to ensure that
they compile and execute to the published design specifications.  The library also performs extensive parameter error checking
and will give verbose error messages when possible.

Essentially the library saves the time of having to implement the ciphers, hashes, prngs yourself.  Typically implementing
useful cryptography is an error prone business which means anything that can save considerable time and effort is a good
thing.

\subsection{What the library IS for?}

The library typically serves as a basis for other protocols and message formats.  For example, it should be possible to 
take the RSA routines out of this library, apply the appropriate message padding and get PKCS compliant RSA routines.  
Similarly SSL protocols could be formed on top  of the low-level symmetric cipher functions.  The goal of this package is 
to provide these low level core functions in a robust and easy to use fashion.

The library also serves well as a toolkit for applications where they don't need to be OpenPGP, PKCS, etc. compliant.
Included are fully operational public key routines for encryption, decryption, signature generation and verification.  
These routines are fully portable but are not conformant to any known set of standards\footnote{With the exception of 
the RSA code which is based on the PKCS \#1 standards.}.  They are all based on established
number theory and cryptography.  

\subsection{What the library IS NOT for?}

The library is not designed to be in anyway an implementation of the SSL or OpenPGP standards.  The library 
is not designed to be compliant with any known form of API or programming hierarchy.  It is not a port of any other 
library and it is not platform specific (like the MS CSP).  So if you're looking to drop in some buzzword 
compliant crypto library this is not for you.  The library has been written from scratch to provide basic functions as 
well as non-standard higher level functions.  

This is not to say that the library is a ``homebrew'' project.  All of the symmetric ciphers and one-way hash functions
conform to published test vectors.  The public key functions are derived from publicly available material and the majority
of the code has been reviewed by a growing community of developers.

\subsubsection{Why not?}
You may be asking why I didn't choose to go all out and support standards like P1363, PKCS and the whole lot.  The reason
is quite simple too much money gets in the way.  When I tried to access the P1363 draft documents and was denied (it 
requires a password) I realized that they're just a business anyways.  See what happens is a company will sit down and
invent a ``standard''.  Then they try to sell it to as many people as they can.  All of a sudden this ``standard'' is 
everywhere.  Then the standard is updated every so often to keep people dependent.  Then you become RSA.  If people are 
supposed to support these standards they had better make them more accessible.

\section{Why did I write it?}
You may be wondering, ``Tom, why did you write a crypto library.  I already have one.''.  Well the reason falls into
two categories:
\begin{enumerate}
    \item I am too lazy to figure out someone else's API.  I'd rather invent my own simpler API and use that.
    \item It was (still is) good coding practice.
\end{enumerate}

The idea is that I am not striving to replace OpenSSL or Crypto++ or Cryptlib or etc.  I'm trying to write my 
{\bf own} crypto library and hopefully along the way others will appreciate the work.

With this library all core functions (ciphers, hashes, prngs) have the {\bf exact} same prototype definition.  They all load
and store data in a format independent of the platform.  This means if you encrypt with Blowfish on a PPC it should decrypt
on an x86 with zero problems.  The consistent API also means that if you learn how to use blowfish with my library you 
know how to use Safer+ or RC6 or Serpent or ... as well.  With all of the core functions there are central descriptor tables 
that can be used to make a program automatically pick between ciphers, hashes and PRNGs at runtime.  That means your 
application can support all ciphers/hashes/prngs without changing the source code.

\subsection{Modular}
The LibTomCrypt package has also been written to be very modular.  The block ciphers, one-way hashes and
pseudo-random number generators (PRNG) are all used within the API through ``descriptor'' tables which 
are essentially structures with pointers to functions.  While you can still call particular functions
directly (\textit{e.g. sha256\_process()}) this descriptor interface allows the developer to customize their
usage of the library.

For example, consider a hardware platform with a specialized RNG device.  Obviously one would like to tap
that for the PRNG needs within the library (\textit{e.g. making a RSA key}).  All the developer has todo
is write a descriptor and the few support routines required for the device.  After that the rest of the 
API can make use of it without change.  Similiarly imagine a few years down the road when AES2 (\textit{or whatever they call it}) is
invented.  It can be added to the library and used within applications with zero modifications to the
end applications provided they are written properly.

This flexibility within the library means it can be used with any combination of primitive algorithms and 
unlike libraries like OpenSSL is not tied to direct routines.  For instance, in OpenSSL there are CBC block
mode routines for every single cipher.  That means every time you add or remove a cipher from the library
you have to update the associated support code as well.  In LibTomCrypt the associated code (\textit{chaining modes in this case})
are not directly tied to the ciphers.  That is a new cipher can be added to the library by simply providing 
the key setup, ECB decrypt and encrypt and test vector routines.  After that all five chaining mode routines
can make use of the cipher right away.


\section{License}

All of the source code except for the following files have been written by the author or donated to the project
under a public domain license:

\begin{enumerate}
   \item rc2.c
   \item safer.c
\end{enumerate}

`mpi.c'' was originally written by Michael Fromberger (sting@linguist.dartmouth.edu) but has since been replaced with my LibTomMath
library.

``rc2.c'' is based on publicly available code that is not attributed to a person from the given source.  ``safer.c''
was written by Richard De Moliner (demoliner@isi.ee.ethz.ch) and seems to be free for use.

The project is hereby released as public domain.

\section{Patent Disclosure}

The author (Tom St Denis) is not a patent lawyer so this section is not to be treated as legal advice.  To the best
of the authors knowledge the only patent related issues within the library are the RC5 and RC6 symmetric block ciphers.  
They can be removed from a build by simply commenting out the two appropriate lines in ``mycrypt\_custom.h''.  The rest
of the ciphers and hashes are patent free or under patents that have since expired.

The RC2 and RC4 symmetric ciphers are not under patents but are under trademark regulations.  This means you can use 
the ciphers you just can't advertise that you are doing so.  

\section{Thanks}
I would like to give thanks to the following people (in no particular order) for helping me develop this project from
early on:
\begin{enumerate}
   \item Richard van de Laarschot
   \item Richard Heathfield
   \item Ajay K. Agrawal
   \item Brian Gladman
   \item Svante Seleborg
   \item Clay Culver
   \item Jason Klapste
   \item Dobes Vandermeer
   \item Daniel Richards
   \item Wayne Scott
   \item Andrew Tyler
   \item Sky Schulz
   \item Christopher Imes
\end{enumerate}

There have been quite a few other people as well.  Please check the change log to see who else has contributed from
time to time.


\chapter{The Application Programming Interface (API)}
\section{Introduction}
\index{CRYPT\_ERROR} \index{CRYPT\_OK}

In general the API is very simple to memorize and use.  Most of the functions return either {\bf void} or {\bf int}.  Functions
that return {\bf int} will return {\bf CRYPT\_OK} if the function was successful or one of the many error codes 
if it failed.  Certain functions that return int will return $-1$ to indicate an error.  These functions will be explicitly
commented upon.  When a function does return a CRYPT error code it can be translated into a string with

\index{error\_to\_string()}
\begin{verbatim}
const char *error_to_string(int err);
\end{verbatim}

An example of handling an error is:
\begin{verbatim}
void somefunc(void)
{
   int err;
   
   /* call a cryptographic function */
   if ((err = some_crypto_function(...)) != CRYPT_OK) {
      printf("A crypto error occured, %s\n", error_to_string(err));
      /* perform error handling */
   }
   /* continue on if no error occured */
}
\end{verbatim}

There is no initialization routine for the library and for the most part the code is thread safe.  The only thread
related issue is if you use the same symmetric cipher, hash or public key state data in multiple threads.  Normally
that is not an issue.

To include the prototypes for ``LibTomCrypt.a'' into your own program simply include ``mycrypt.h'' like so:
\begin{verbatim}
#include <mycrypt.h>
int main(void) {
    return 0;
}
\end{verbatim}

The header file ``mycrypt.h'' also includes ``stdio.h'', ``string.h'', ``stdlib.h'', ``time.h'', ``ctype.h'' and ``mpi.h''
(the bignum library routines).

\section{Macros}

There are a few helper macros to make the coding process a bit easier.  The first set are related to loading and storing
32/64-bit words in little/big endian format.  The macros are:

\index{STORE32L} \index{STORE64L} \index{LOAD32L} \index{LOAD64L}
\index{STORE32H} \index{STORE64H} \index{LOAD32H} \index{LOAD64H} \index{BSWAP}
\begin{small}
\begin{center}
\begin{tabular}{|c|c|c|}
     \hline STORE32L(x, y) & {\bf unsigned long} x, {\bf unsigned char} *y & $x \to y[0 \ldots 3]$ \\
     \hline STORE64L(x, y) & {\bf unsigned long long} x, {\bf unsigned char} *y & $x \to y[0 \ldots 7]$ \\
     \hline LOAD32L(x, y) & {\bf unsigned long} x, {\bf unsigned char} *y & $y[0 \ldots 3] \to x$ \\
     \hline LOAD64L(x, y) & {\bf unsigned long long} x, {\bf unsigned char} *y & $y[0 \ldots 7] \to x$ \\
     \hline STORE32H(x, y) & {\bf unsigned long} x, {\bf unsigned char} *y & $x \to y[3 \ldots 0]$ \\
     \hline STORE64H(x, y) & {\bf unsigned long long} x, {\bf unsigned char} *y & $x \to y[7 \ldots 0]$ \\
     \hline LOAD32H(x, y) & {\bf unsigned long} x, {\bf unsigned char} *y & $y[3 \ldots 0] \to x$ \\
     \hline LOAD64H(x, y) & {\bf unsigned long long} x, {\bf unsigned char} *y & $y[7 \ldots 0] \to x$ \\
     \hline BSWAP(x) & {\bf unsigned long} x & Swaps the byte order of x. \\
     \hline
\end{tabular}
\end{center}
\end{small}

There are 32-bit cyclic rotations as well:
\index{ROL} \index{ROR}
\begin{center}
\begin{tabular}{|c|c|c|}
     \hline ROL(x, y) & {\bf unsigned long} x, {\bf unsigned long} y & $x << y$ \\
     \hline ROR(x, y) & {\bf unsigned long} x, {\bf unsigned long} y & $x >> y$ \\
     \hline
\end{tabular}
\end{center}

\section{Functions with Variable Length Output}
Certain functions such as (for example) ``rsa\_export()'' give an output that is variable length.  To prevent buffer overflows you
must pass it the length of the buffer\footnote{Extensive error checking is not in place but it will be in future releases so it is a good idea to follow through with these guidelines.} where
the output will be stored.  For example:
\begin{small}
\begin{verbatim}
#include <mycrypt.h>
int main(void) {
    rsa_key key;
    unsigned char buffer[1024];
    unsigned long x;
    int err;

    /* ... Make up the RSA key somehow */

    /* lets export the key, set x to the size of the output buffer */
    x = sizeof(buffer);
    if ((err = rsa_export(buffer, &x, PK_PUBLIC, &key)) != CRYPT_OK) {
       printf("Export error: %s\n", error_to_string(err));
       return -1;
    }
    
    /* if rsa_export() was successful then x will have the size of the output */
    printf("RSA exported key takes %d bytes\n", x);

    /* ... do something with the buffer */

    return 0;
}
\end{verbatim}
\end{small}
In the above example if the size of the RSA public key was more than 1024 bytes this function would not store anything in
either ``buffer'' or ``x'' and simply return an error code.  If the function suceeds it stores the length of the output
back into ``x'' so that the calling application will know how many bytes used.

\section{Functions that need a PRNG}
Certain functions such as ``rsa\_make\_key()'' require a PRNG.  These functions do not setup the PRNG themselves so it is 
the responsibility of the calling function to initialize the PRNG before calling them.

\section{Functions that use Arrays of Octets}
Most functions require inputs that are arrays of the data type ``unsigned char''.  Whether it is a symmetric key, IV
for a chaining mode or public key packet it is assumed that regardless of the actual size of ``unsigned char'' only the
lower eight bits contain data.  For example, if you want to pass a 256 bit key to a symmetric ciphers setup routine
you must pass it in (a pointer to) an array of 32 ``unsigned char'' variables.  Certain routines 
(such as SAFER+) take special care to work properly on platforms where an ``unsigned char'' is not eight bits.

For the purposes of this library the term ``byte'' will refer to an octet or eight bit word.  Typically an array of
type ``byte'' will be synonymous with an array of type ``unsigned char''.

\chapter{Symmetric Block Ciphers}
\section{Core Functions}

Libtomcrypt provides several block ciphers all in a plain vanilla ECB block mode.  Its important to first note that you 
should never use the ECB modes directly to encrypt data.  Instead you should use the ECB functions to make a chaining mode
or use one of the provided chaining modes.  All of the ciphers are written as ECB interfaces since it allows the rest of
the API to grow in a modular fashion.

All ciphers store their scheduled keys in a single data type called ``symmetric\_key''.  This allows all ciphers to 
have the same prototype and store their keys as  naturally as possible.  All ciphers provide five visible functions which
are (given that XXX is the name of the cipher):
\index{Cipher Setup}
\begin{verbatim}
int XXX_setup(const unsigned char *key, int keylen, int rounds,
              symmetric_key *skey);
\end{verbatim}

The XXX\_setup() routine will setup the cipher to be used with a given number of rounds and a given key length (in bytes).
The number of rounds can be set to zero to use the default, which is generally a good idea.

If the function returns successfully the variable ``skey'' will have a scheduled key stored in it.  Its important to note
that you should only used this scheduled key with the intended cipher.  For example, if you call 
``blowfish\_setup()'' do not pass the scheduled key onto ``rc5\_ecb\_encrypt()''.  All setup functions do not allocate 
memory off the heap so when you are done with a key you can simply discard it (e.g. they can be on the stack).

To encrypt or decrypt a block in ECB mode there are these two functions:
\index{Cipher Encrypt} \index{Cipher Decrypt}
\begin{verbatim}
void XXX_ecb_encrypt(const unsigned char *pt, unsigned char *ct,
                     symmetric_key *skey);

void XXX_ecb_decrypt(const unsigned char *ct, unsigned char *pt,
                     symmetric_key *skey);
\end{verbatim}
These two functions will encrypt or decrypt (respectively) a single block of text\footnote{The size of which depends on
which cipher you are using.} and store the result where you want it.  It is possible that the input and output buffer are 
the same buffer.  For the encrypt function ``pt''\footnote{pt stands for plaintext.} is the input and ``ct'' is the output.
For the decryption function its the opposite.  To test a particular cipher against test vectors\footnote{As published in their design papers.} call: \index{Cipher Testing}
\begin{verbatim}
int XXX_test(void);
\end{verbatim}
This function will return {\bf CRYPT\_OK} if the cipher matches the test vectors from the design publication it is 
based upon.  Finally for each cipher there is a function which will help find a desired key size:
\begin{verbatim}
int XXX_keysize(int *keysize);
\end{verbatim}
Essentially it will round the input keysize in ``keysize'' down to the next appropriate key size.  This function
return {\bf CRYPT\_OK} if the key size specified is acceptable.  For example:
\begin{small}
\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   int keysize, err;

   /* now given a 20 byte key what keysize does Twofish want to use? */
   keysize = 20;
   if ((err = twofish_keysize(&keysize)) != CRYPT_OK) {
      printf("Error getting key size: %s\n", error_to_string(err));
      return -1;
   }
   printf("Twofish suggested a key size of %d\n", keysize);
   return 0;
}
\end{verbatim}
\end{small}
This should indicate a keysize of sixteen bytes is suggested.  An example snippet that encodes a block with 
Blowfish in ECB mode is below.

\begin{small}
\begin{verbatim}
#include <mycrypt.h>
int main(void)
{ 
   unsigned char pt[8], ct[8], key[8];
   symmetric_key skey;
   int err;

   /* ... key is loaded appropriately in ``key'' ... */
   /* ... load a block of plaintext in ``pt'' ... */

   /* schedule the key */
   if ((err = blowfish_setup(key,     /* the key we will use */
                               8,     /* key is 8 bytes (64-bits) long */
                               0,     /* 0 == use default # of rounds */
                           &skey)     /* where to put the scheduled key */
       ) != CRYPT_OK) {
      printf("Setup error: %s\n", error_to_string(err));
      return -1;
   }

   /* encrypt the block */
   blowfish_ecb_encrypt(pt,             /* encrypt this 8-byte array */
                        ct,             /* store encrypted data here */ 
                        &skey);         /* our previously scheduled key */

   /* decrypt the block */
   blowfish_ecb_decrypt(ct,             /* decrypt this 8-byte array */
                        pt,             /* store decrypted data here */
                        &skey);         /* our previously scheduled key */

   return 0;
}
\end{verbatim}
\end{small}

\section{Key Sizes and Number of Rounds}
\index{Symmetric Keys}
As a general rule of thumb do not use symmetric keys under 80 bits if you can.  Only a few of the ciphers support smaller
keys (mainly for test vectors anyways).  Ideally your application should be making at least 256 bit keys.  This is not
because you're supposed to be paranoid.  Its because if your PRNG has a bias of any sort the more bits the better.  For
example, if you have $\mbox{Pr}\left[X = 1\right] = {1 \over 2} \pm \gamma$ where $\vert \gamma \vert > 0$ then the
total amount of entropy in N bits is $N \cdot -log_2\left ({1 \over 2} + \vert \gamma \vert \right)$.  So if $\gamma$
were $0.25$ (a severe bias) a 256-bit string would have about 106 bits of entropy whereas a 128-bit string would have
only 53 bits of entropy.

The number of rounds of most ciphers is not an option you can change.  Only RC5 allows you to change the number of
rounds.  By passing zero as the number of rounds all ciphers will use their default number of rounds.  Generally the
ciphers are configured such that the default number of rounds provide adequate security for the given block size.

\section{The Cipher Descriptors}
\index{Cipher Descriptor}
To facilitate automatic routines an array of cipher descriptors is provided in the array ``cipher\_descriptor''.  An element
of this array has the following format:

\begin{verbatim}
struct _cipher_descriptor {
   char *name;
   unsigned long min_key_length, max_key_length, 
                 block_length, default_rounds;
   int  (*setup)      (const unsigned char *key, int keylength, 
                       int num_rounds, symmetric_key *skey);
   void (*ecb_encrypt)(const unsigned char *pt, unsigned char *ct, 
                       symmetric_key *key);
   void (*ecb_decrypt)(const unsigned char *ct, unsigned char *pt,
                       symmetric_key *key);
   int  (*test)       (void);
   int  (*keysize)    (int *desired_keysize);
};
\end{verbatim}

Where ``name'' is the lower case ASCII version of the name.  The fields ``min\_key\_length'', ``max\_key\_length'' and
``block\_length'' are all the number of bytes not bits.  As a good rule of thumb it is assumed that the cipher supports
the min and max key lengths but not always everything in between.  The ``default\_rounds'' field is the default number
of rounds that will be used.

The remaining fields are all pointers to the core functions for each cipher.  The end of the cipher\_descriptor array is
marked when ``name'' equals {\bf NULL}.

As of this release the current cipher\_descriptors elements are

\index{Cipher descriptor table}
\begin{small}
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
     \hline Name & Descriptor Name & Block Size & Key Range & Rounds \\
     \hline Blowfish & blowfish\_desc & 8 & 8 $\ldots$ 56 & 16 \\
     \hline X-Tea & xtea\_desc & 8 & 16 & 32 \\
     \hline RC2 & rc2\_desc & 8 & 8 $\ldots$ 128 & 16 \\
     \hline RC5-32/12/b & rc5\_desc & 8 & 8 $\ldots$ 128 & 12 $\ldots$ 24 \\
     \hline RC6-32/20/b & rc6\_desc & 16 & 8 $\ldots$ 128 & 20 \\
     \hline SAFER+ & saferp\_desc &16 & 16, 24, 32 & 8, 12, 16 \\
     \hline Safer K64   & safer\_k64\_desc & 8 & 8 & 6 $\ldots$ 13 \\
     \hline Safer SK64  & safer\_sk64\_desc & 8 & 8 & 6 $\ldots$ 13 \\
     \hline Safer K128  & safer\_k128\_desc & 8 & 16 & 6 $\ldots$ 13 \\
     \hline Safer SK128 & safer\_sk128\_desc & 8 & 16 & 6 $\ldots$ 13 \\
     \hline AES & aes\_desc & 16 & 16, 24, 32 & 10, 12, 14 \\
                & aes\_enc\_desc & 16 & 16, 24, 32 & 10, 12, 14 \\
     \hline Twofish & twofish\_desc & 16 & 16, 24, 32 & 16 \\
     \hline DES & des\_desc & 8 & 7 & 16 \\
     \hline 3DES (EDE mode) & des3\_desc & 8 & 21 & 16 \\
     \hline CAST5 (CAST-128) & cast5\_desc & 8 & 5 $\ldots$ 16 & 12, 16 \\
     \hline Noekeon & noekeon\_desc & 16 & 16 & 16 \\
     \hline Skipjack & skipjack\_desc & 8 & 10 & 32 \\
     \hline
\end{tabular}
\end{center}
\end{small}

\subsection{Notes}
\begin{small}
\begin{enumerate}
\item
For AES (also known as Rijndael) there are four descriptors which complicate issues a little.  The descriptors 
rijndael\_desc and rijndael\_enc\_desc provide the cipher named ``rijndael''.  The descriptors aes\_desc and 
aes\_enc\_desc provide the cipher name ``aes''.  Functionally both ``rijndael'' and ``aes'' are the same cipher.  The
only difference is when you call find\_cipher() you have to pass the correct name.  The cipher descriptors with ``enc'' 
in the middle (e.g. rijndael\_enc\_desc) are related to an implementation of Rijndael with only the encryption routine
and tables.  The decryption and self--test function pointers of both ``encrypt only'' descriptors are set to \textbf{NULL} and 
should not be called.

The ``encrypt only'' descriptors are useful for applications that only use the encryption function of the cipher.  Algorithms such
as EAX, PMAC and OMAC only require the encryption function.  So far this ``encrypt only'' functionality has only been implemented for
Rijndael as it makes the most sense for this cipher.

\item
For the 64-bit SAFER famliy of ciphers (e.g K64, SK64, K128, SK128) the ecb\_encrypt() and ecb\_decrypt()
functions are the same.  So if you want to use those functions directly just call safer\_ecb\_encrypt()
or safer\_ecb\_decrypt() respectively.

\item
Note that for ``DES'' and ``3DES'' they use 8 and 24 byte keys but only 7 and 21 [respectively] bytes of the keys are in
fact used for the purposes of encryption.  My suggestion is just to use random 8/24 byte keys instead of trying to make a 8/24
byte string from the real 7/21 byte key.

\item
Note that ``Twofish'' has additional configuration options that take place at build time.  These options are found in
the file ``mycrypt\_cfg.h''.  The first option is ``TWOFISH\_SMALL'' which when defined will force the Twofish code
to not pre-compute the Twofish ``$g(X)$'' function as a set of four $8 \times 32$ s-boxes.  This means that a scheduled
key will require less ram but the resulting cipher will be slower.  The second option is ``TWOFISH\_TABLES'' which when
defined will force the Twofish code to use pre-computed tables for the two s-boxes $q_0, q_1$ as well as the multiplication
by the polynomials 5B and EF used in the MDS multiplication.  As a result the code is faster and slightly larger.  The
speed increase is useful when ``TWOFISH\_SMALL'' is defined since the s-boxes and MDS multiply form the heart of the
Twofish round function.

\index{Twofish build options}
\begin{small}
\begin{center}
\begin{tabular}{|l|l|l|}
\hline TWOFISH\_SMALL & TWOFISH\_TABLES & Speed and Memory (per key) \\
\hline undefined & undefined & Very fast, 4.2KB of ram. \\
\hline undefined & defined & Faster keysetup, larger code. \\
\hline defined & undefined & Very slow, 0.2KB of ram. \\
\hline defined & defined & Faster, 0.2KB of ram, larger code. \\
\hline
\end{tabular}
\end{center}
\end{small}

\end{enumerate}
\end{small}

To work with the cipher\_descriptor array there is a function:
\index{find\_cipher()}
\begin{verbatim}
int find_cipher(char *name)
\end{verbatim}
Which will search for a given name in the array.  It returns negative one if the cipher is not found, otherwise it returns
the location in the array where the cipher was found.  For example, to indirectly setup Blowfish you can also use:
\begin{small}
\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   unsigned char key[8];
   symmetric_key skey;
   int err;

   /* you must register a cipher before you use it */
   if (register_cipher(&blowfish_desc)) == -1) {
      printf("Unable to register Blowfish cipher.");
      return -1;
   }

   /* generic call to function (assuming the key in key[] was already setup) */
   if ((err = cipher_descriptor[find_cipher("blowfish")].setup(key, 8, 0, &skey)) != 
       CRYPT_OK) {
      printf("Error setting up Blowfish: %s\n", error_to_string(err));
      return -1;
   }

   /* ... use cipher ... */
}
\end{verbatim}
\end{small}

A good safety would be to check the return value of ``find\_cipher()'' before accessing the desired function.  In order
to use a cipher with the descriptor table you must register it first using:
\index{register\_cipher()}
\begin{verbatim}
int register_cipher(const struct _cipher_descriptor *cipher);
\end{verbatim}
Which accepts a pointer to a descriptor and returns the index into the global descriptor table.  If an error occurs such
as there is no more room (it can have 32 ciphers at most) it will return {\bf{-1}}.  If you try to add the same cipher more
than once it will just return the index of the first copy.  To remove a cipher call:
\index{unregister\_cipher()}
\begin{verbatim}
int unregister_cipher(const struct _cipher_descriptor *cipher);
\end{verbatim}
Which returns {\bf CRYPT\_OK} if it removes it otherwise it returns {\bf CRYPT\_ERROR}.  Consider:
\begin{small}
\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   int err;
   
   /* register the cipher */
   if (register_cipher(&rijndael_desc) == -1) {
      printf("Error registering Rijndael\n");
      return -1;
   }

   /* use Rijndael */

   /* remove it */
   if ((err = unregister_cipher(&rijndael_desc)) != CRYPT_OK) {
      printf("Error removing Rijndael: %s\n", error_to_string(err));
      return -1;
   }

   return 0;
}
\end{verbatim}
\end{small}
This snippet is a small program that registers only Rijndael only.  

\section{Symmetric Modes of Operations}
\subsection{Background}
A typical symmetric block cipher can be used in chaining modes to effectively encrypt messages larger than the block
size of the cipher.  Given a key $k$, a plaintext $P$ and a cipher $E$ we shall denote the encryption of the block
$P$ under the key $k$ as $E_k(P)$.  In some modes there exists an initial vector denoted as $C_{-1}$.

\subsubsection{ECB Mode}
\index{ECB mode}
ECB or Electronic Codebook Mode is the simplest method to use.  It is given as:
\begin{equation}
C_i = E_k(P_i)
\end{equation}
This mode is very weak since it allows people to swap blocks and perform replay attacks if the same key is used more
than once.

\subsubsection{CBC Mode}
\index{CBC mode}
CBC or Cipher Block Chaining mode is a simple mode designed to prevent trivial forms of replay and swap attacks on ciphers.
It is given as:
\begin{equation}
C_i = E_k(P_i \oplus C_{i - 1})
\end{equation}
It is important that the initial vector be unique and preferably random for each message encrypted under the same key.

\subsubsection{CTR Mode}
\index{CTR mode}
CTR or Counter Mode is a mode which only uses the encryption function of the cipher.  Given a initial vector which is
treated as a large binary counter the CTR mode is given as:
\begin{eqnarray}
C_{-1} = C_{-1} + 1\mbox{ }(\mbox{mod }2^W) \nonumber \\
C_i = P_i \oplus E_k(C_{-1})
\end{eqnarray}
Where $W$ is the size of a block in bits (e.g. 64 for Blowfish).  As long as the initial vector is random for each message
encrypted under the same key replay and swap attacks are infeasible.  CTR mode may look simple but it is as secure
as the block cipher is under a chosen plaintext attack (provided the initial vector is unique).

\subsubsection{CFB Mode}
\index{CFB mode}
CFB or Ciphertext Feedback Mode is a mode akin to CBC.  It is given as:
\begin{eqnarray}
C_i = P_i \oplus C_{-1} \nonumber \\
C_{-1} = E_k(C_i)
\end{eqnarray}
Note that in this library the output feedback width is equal to the size of the block cipher.  That is this mode is used
to encrypt whole blocks at a time.  However, the library will buffer data allowing the user to encrypt or decrypt partial
blocks without a delay.  When this mode is first setup it will initially encrypt the initial vector as required.

\subsubsection{OFB Mode}
\index{OFB mode}
OFB or Output Feedback Mode is a mode akin to CBC as well.  It is given as:
\begin{eqnarray}
C_{-1} = E_k(C_{-1}) \nonumber \\
C_i = P_i \oplus C_{-1}
\end{eqnarray}
Like the CFB mode the output width in CFB mode is the same as the width of the block cipher.  OFB mode will also
buffer the output which will allow you to encrypt or decrypt partial blocks without delay.

\subsection{Choice of Mode}
My personal preference is for the CTR mode since it has several key benefits:
\begin{enumerate}
   \item No short cycles which is possible in the OFB and CFB modes.
   \item Provably as secure as the block cipher being used under a chosen plaintext attack.
   \item Technically does not require the decryption routine of the cipher.
   \item Allows random access to the plaintext.
   \item Allows the encryption of block sizes that are not equal to the size of the block cipher.
\end{enumerate}
The CTR, CFB and OFB routines provided allow you to encrypt block sizes that differ from the ciphers block size.  They 
accomplish this by buffering the data required to complete a block.  This allows you to encrypt or decrypt any size 
block of memory with either of the three modes.

The ECB and CBC modes process blocks of the same size as the cipher at a time.  Therefore they are less flexible than the
other modes.

\subsection{Implementation}
\index{CBC Mode} \index{CTR Mode}
\index{OFB Mode} \index{CFB Mode}
The library provides simple support routines for handling CBC, CTR, CFB, OFB and ECB encoded messages.  Assuming the mode 
you want is XXX there is a structure called ``symmetric\_XXX'' that will contain the information required to
use that mode.  They have identical setup routines (except ECB mode for obvious reasons):
\index{ecb\_start()} \index{cfb\_start()} \index{cbc\_start()} \index{ofb\_start()} \index{ctr\_start()}
\begin{verbatim}
int XXX_start(int cipher, const unsigned char *IV, 
              const unsigned char *key, int keylen, 
              int num_rounds, symmetric_XXX *XXX);

int ecb_start(int cipher, const unsigned char *key, int keylen, 
              int num_rounds, symmetric_ECB *ecb);
\end{verbatim}

In each case ``cipher'' is the index into the cipher\_descriptor array of the cipher you want to use.  The ``IV'' value is 
the initialization vector to be used with the cipher.  You must fill the IV yourself and it is assumed they are the same 
length as the block size\footnote{In otherwords the size of a block of plaintext for the cipher, e.g. 8 for DES, 16 for AES, etc.} 
of the cipher you choose.  It is important that the IV  be random for each unique message you want to encrypt.  The 
parameters ``key'', ``keylen'' and ``num\_rounds'' are the same as in the XXX\_setup() function call.  The final parameter 
is a pointer to the structure you want to hold the information for the mode of operation.

Both routines return {\bf CRYPT\_OK} if the cipher initialized correctly, otherwise they return an error code.  To 
actually encrypt or decrypt the following routines are provided:
\index{ecb\_encrypt()} \index{ecb\_decrypt()} \index{cfb\_encrypt()} \index{cfb\_decrypt()} 
\index{cbc\_encrypt()} \index{cbc\_decrypt()} \index{ofb\_encrypt()} \index{ofb\_decrypt()} \index{ctr\_encrypt()} \index{ctr\_decrypt()}
\begin{verbatim}
int XXX_encrypt(const unsigned char *pt, unsigned char *ct, 
                symmetric_XXX *XXX);
int XXX_decrypt(const unsigned char *ct, unsigned char *pt,
                symmetric_XXX *XXX);

int YYY_encrypt(const unsigned char *pt, unsigned char *ct, 
                unsigned long len, symmetric_YYY *YYY);
int YYY_decrypt(const unsigned char *ct, unsigned char *pt, 
                unsigned long len, symmetric_YYY *YYY);
\end{verbatim}
Where ``XXX'' is one of (ecb, cbc) and ``YYY'' is one of (ctr, ofb, cfb).  In the CTR, OFB and CFB cases ``len'' is the
size of the buffer (as number of chars) to encrypt or decrypt.  The CTR, OFB and CFB modes are order sensitive but not
chunk sensitive.  That is you can encrypt ``ABCDEF'' in three calls like ``AB'', ``CD'', ``EF'' or two like ``ABCDE'' and ``F''
and end up with the same ciphertext.  However, encrypting ``ABC'' and ``DABC'' will result in different ciphertexts.  All
five of the modes will return {\bf CRYPT\_OK} on success from the encrypt or decrypt functions.

To decrypt in either mode you simply perform the setup like before (recall you have to fetch the IV value you used)
and use the decrypt routine on all of the blocks.

To change or read the IV of a previously initialized chaining mode use the following two functions.

\index{cbc\_setiv()} \index{cbc\_getiv()} \index{ofb\_setiv()} \index{ofb\_getiv()} \index{cfb\_setiv()} \index{cfb\_getiv()}
\index{ctr\_setiv()} \index{ctr\_getiv()}
\begin{verbatim}
int XXX_getiv(unsigned char *IV, unsigned long *len, symmetric_XXX *XXX);
int XXX_setiv(const unsigned char *IV, unsigned long len, symmetric_XXX *XXX);
\end{verbatim}

The XXX\_getiv function will read the IV out of the chaining mode and store it into ``IV'' along with the length of the IV 
stored in ``len''.  The XXX\_setiv will initialize the chaining mode state as if the original IV were the new IV specified.  The length
of the IV passed in must be the size of the ciphers block size.

The XXX\_setiv functions are handy if you wish to change the IV without re--keying the cipher.  

\newpage
\begin{small}
\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   unsigned char key[16], IV[16], buffer[512];
   symmetric_CTR ctr;
   int x, err;

   /* register twofish first */
   if (register_cipher(&twofish_desc) == -1) {
      printf("Error registering cipher.\n");
      return -1;
   }

   /* somehow fill out key and IV */

   /* start up CTR mode */
   if ((err = ctr_start(
        find_cipher("twofish"), /* index of desired cipher */
                            IV, /* the initial vector */
                           key, /* the secret key */
                            16, /* length of secret key (16 bytes, 128 bits) */
                             0, /* 0 == default # of rounds */
                         &ctr)  /* where to store initialized CTR state */
      ) != CRYPT_OK) {
      printf("ctr_start error: %s\n", error_to_string(err));
      return -1;
   }

   /* somehow fill buffer than encrypt it */
   if ((err = ctr_encrypt(        buffer, /* plaintext */
                                  buffer, /* ciphertext */
                          sizeof(buffer), /* length of data to encrypt */
                                   &ctr)  /* previously initialized CTR state */
      ) != CRYPT_OK) {
      printf("ctr_encrypt error: %s\n", error_to_string(err));
      return -1;
   }

   /* make use of ciphertext... */

   /* now we want to decrypt so let's use ctr_setiv */
   if ((err = ctr_setiv(  IV, /* the initial IV we gave to ctr_start */
                          16, /* the IV is 16 bytes long */
                        &ctr) /* the ctr state we wish to modify */
       ) != CRYPT_OK) {
      printf("ctr_setiv error: %s\n", error_to_string(err));
      return -1;
   }

   if ((err = ctr_decrypt(        buffer, /* ciphertext */
                                  buffer, /* plaintext */
                          sizeof(buffer), /* length of data to encrypt */
                                   &ctr)  /* previously initialized CTR state */
      ) != CRYPT_OK) {
      printf("ctr_decrypt error: %s\n", error_to_string(err));
      return -1;
   }

   /* clear up and return */
   zeromem(key, sizeof(key));
   zeromem(&ctr, sizeof(ctr));

   return 0;
}
\end{verbatim}
\end{small}

\section{Encrypt and Authenticate Modes}

\subsection{EAX Mode}
LibTomCrypt provides support for a mode called EAX\footnote{See 
M. Bellare, P. Rogaway, D. Wagner, A Conventional Authenticated-Encryption Mode.} in a manner similar to the
way it was intended to be used by the designers.  First a short description of what EAX mode is before I explain how to use it.  
EAX is a mode that requires a cipher, CTR and OMAC support and provides encryption and authentication\footnote{Note that since EAX only requires OMAC and CTR you may use ``encrypt only'' cipher descriptors with this mode.}.  
It is initialized with a random ``nonce'' that can be shared publicly as well as a ``header'' which can be fixed and public as well as a random 
secret symmetric key.

The ``header'' data is meant to be meta-data associated with a stream that isn't private (e.g. protocol messages).  It can
be added at anytime during an EAX stream and is part of the authentication tag.  That is, changes in the meta-data can
be detected by changes in the output tag.

The mode can then process plaintext producing ciphertext as well as compute a partial checksum.  The actual checksum
called a ``tag'' is only emitted when the message is finished.  In the interim though the user can process any arbitrary
sized message block to send to the recipient as ciphertext.  This makes the EAX mode especially suited for streaming modes
of operation.

The mode is initialized with the following function.
\index{eax\_init()}
\begin{verbatim}
int eax_init(eax_state *eax, int cipher, 
             const unsigned char *key, unsigned long keylen,
             const unsigned char *nonce, unsigned long noncelen,
             const unsigned char *header, unsigned long headerlen);
\end{verbatim}

Where ``eax'' is the EAX state.  ``cipher'' is the index of the desired cipher in the descriptor table.  
``key'' is the shared secret symmetric key of length ``keylen''.  ``nonce'' is the random public string of
length ``noncelen''.  ``header'' is the random (or fixed or \textbf{NULL}) header for the message of length
``headerlen''.

When this function completes ``eax'' will be initialized such that you can now either have data decrypted or 
encrypted in EAX mode.  Note that if ``headerlen'' is zero you may pass ``header'' as \textbf{NULL} to indicate
there is no initial header data.

To encrypt or decrypt data in a streaming mode use the following.
\index{eax\_encrypt()} \index{eax\_decrypt()}
\begin{verbatim}
int eax_encrypt(eax_state *eax, const unsigned char *pt, 
                unsigned char *ct, unsigned long length);

int eax_decrypt(eax_state *eax, const unsigned char *ct, 
                unsigned char *pt, unsigned long length);
\end{verbatim}
The function ``eax\_encrypt'' will encrypt the bytes in ``pt'' of ``length'' bytes and store the ciphertext in
``ct''.  Note that ``ct'' and ``pt'' may be the same region in memory.   This function will also send the ciphertext
through the OMAC function.  The function ``eax\_decrypt'' decrypts ``ct'' and stores it in ``pt''.  This also allows 
``pt'' and ``ct'' to be the same region in memory.  

You cannot both encrypt or decrypt with the same ``eax'' context.  For bi-directional communication you
will need to initialize two EAX contexts (preferably with different headers and nonces).  

Note that both of these functions allow you to send the data in any granularity but the order is important.  While
the eax\_init() function allows you to add initial header data to the stream you can also add header data during the
EAX stream with the following.

\index{eax\_addheader()}
\begin{verbatim}
int eax_addheader(eax_state *eax, 
                  const unsigned char *header, unsigned long length);
\end{verbatim}

This will add the ``length'' bytes from ``header'' to the given ``eax'' stream.  Once the message is finished the 
``tag'' (checksum) may be computed with the following function.

\index{eax\_done()}
\begin{verbatim}
int eax_done(eax_state *eax, 
             unsigned char *tag, unsigned long *taglen);
\end{verbatim}
This will terminate the EAX state ``eax'' and store upto ``taglen'' bytes of the message tag in ``tag''.  The function
then stores how many bytes of the tag were written out back into ``taglen''.

The EAX mode code can be tested to ensure it matches the test vectors by calling the following function.
\index{eax\_test()}
\begin{verbatim}
int eax_test(void);
\end{verbatim}
This requires that the AES (or Rijndael) block cipher be registered with the cipher\_descriptor table first.

\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   int           err;
   eax_state     eax;
   unsigned char pt[64], ct[64], nonce[16], key[16], tag[16];
   unsigned long taglen;

   if (register_cipher(&rijndael_desc) == -1) {
      printf("Error registering Rijndael");
      return EXIT_FAILURE;
   }

   /* ... make up random nonce and key ... */

   /* initialize context */
   if ((err = eax_init(            &eax,  /* the context */
                find_cipher("rijndael"),  /* cipher we want to use */
                                  nonce,  /* our state nonce */
                                     16,  /* none is 16 bytes */
                              "TestApp",  /* example header, identifies this program */
                                      7)  /* length of the header */
       ) != CRYPT_OK) {
      printf("Error eax_init: %s", error_to_string(err));
      return EXIT_FAILURE;
   }

   /* now encrypt data, say in a loop or whatever */
   if ((err = eax_encrypt(     &eax,      /* eax context */
                                 pt,      /* plaintext  (source) */
                                 ct,      /* ciphertext (destination) */
                          sizeof(pt)      /* size of plaintext */
      ) != CRYPT_OK) {
      printf("Error eax_encrypt: %s", error_to_string(err));
      return EXIT_FAILURE;
   }

   /* finish message and get authentication tag */
   taglen = sizeof(tag);
   if ((err = eax_done(   &eax,           /* eax context */
                           tag,           /* where to put tag */
                       &taglen            /* length of tag space */
      ) != CRYPT_OK) {
      printf("Error eax_done: %s", error_to_string(err));
      return EXIT_FAILURE;
   }

   /* now we have the authentication tag in "tag" and it's taglen bytes long */

}
\end{verbatim}
                       
You can also perform an entire EAX state on a block of memory in a single function call with the 
following functions.


\index{eax\_encrypt\_authenticate\_memory} \index{eax\_decrypt\_verify\_memory}
\begin{verbatim}
int eax_encrypt_authenticate_memory(int cipher,
    const unsigned char *key,    unsigned long keylen,
    const unsigned char *nonce,  unsigned long noncelen,
    const unsigned char *header, unsigned long headerlen,
    const unsigned char *pt,     unsigned long ptlen,
          unsigned char *ct,
          unsigned char *tag,    unsigned long *taglen);

int eax_decrypt_verify_memory(int cipher,
    const unsigned char *key,    unsigned long keylen,
    const unsigned char *nonce,  unsigned long noncelen,
    const unsigned char *header, unsigned long headerlen,
    const unsigned char *ct,     unsigned long ctlen,
          unsigned char *pt,
          unsigned char *tag,    unsigned long taglen,
          int           *res);
\end{verbatim}

Both essentially just call eax\_init() followed by eax\_encrypt() (or eax\_decrypt() respectively) and eax\_done().  The parameters
have the same meaning as with those respective functions.  

The only difference is eax\_decrypt\_verify\_memory() does not emit a tag.  Instead you pass it a tag as input and it compares it against
the tag it computed while decrypting the message.  If the tags match then it stores a $1$ in ``res'', otherwise it stores a $0$.

\subsection{OCB Mode}
LibTomCrypt provides support for a mode called OCB\footnote{See 
P. Rogaway, M. Bellare, J. Black, T. Krovetz, ``OCB: A Block Cipher Mode of Operation for Efficient Authenticated Encryption''.}
.  OCB is an encryption protocol that simultaneously provides authentication.  It is slightly faster to use than EAX mode
but is less flexible.  Let's review how to initialize an OCB context.

\index{ocb\_init()}
\begin{verbatim}
int ocb_init(ocb_state *ocb, int cipher, 
             const unsigned char *key, unsigned long keylen, 
             const unsigned char *nonce);
\end{verbatim}

This will initialize the ``ocb'' context using cipher descriptor ``cipher''.  It will use a ``key'' of length ``keylen''
and the random ``nonce''.  Note that ``nonce'' must be a random (public) string the same length as the block ciphers
block size (e.g. 16 bytes for AES).

This mode has no ``Associated Data'' like EAX mode does which means you cannot authenticate metadata along with the stream.
To encrypt or decrypt data use the following.

\index{ocb\_encrypt()} \index{ocb\_decrypt()}
\begin{verbatim}
int ocb_encrypt(ocb_state *ocb, const unsigned char *pt, unsigned char *ct);
int ocb_decrypt(ocb_state *ocb, const unsigned char *ct, unsigned char *pt);
\end{verbatim}

This will encrypt (or decrypt for the latter) a fixed length of data from ``pt'' to ``ct'' (vice versa for the latter).  
They assume that ``pt'' and ``ct'' are the same size as the block cipher's block size.  Note that you cannot call 
both functions given a single ``ocb'' state.  For bi-directional communication you will have to initialize two ``ocb''
states (with different nonces).  Also ``pt'' and ``ct'' may point to the same location in memory.

When you are finished encrypting the message you call the following function to compute the tag.

\index{ocb\_done\_encrypt()}
\begin{verbatim}
int ocb_done_encrypt(ocb_state *ocb, 
                     const unsigned char *pt, unsigned long ptlen,
                           unsigned char *ct, 
                           unsigned char *tag, unsigned long *taglen);
\end{verbatim}

This will terminate an encrypt stream ``ocb''.  If you have trailing bytes of plaintext that will not complete a block 
you can pass them here.  This will also encrypt the ``ptlen'' bytes in ``pt'' and store them in ``ct''.  It will also
store upto ``taglen'' bytes of the tag into ``tag''.

Note that ``ptlen'' must be less than or equal to the block size of block cipher chosen.  Also note that if you have 
an input message equal to the length of the block size then you pass the data here (not to ocb\_encrypt()) only.  

To terminate a decrypt stream and compared the tag you call the following.

\index{ocb\_done\_decrypt()}
\begin{verbatim}
int ocb_done_decrypt(ocb_state *ocb, 
                     const unsigned char *ct,  unsigned long ctlen,
                           unsigned char *pt, 
                     const unsigned char *tag, unsigned long taglen, 
                           int *res);
\end{verbatim}

Similarly to the previous function you can pass trailing message bytes into this function.  This will compute the 
tag of the message (internally) and then compare it against the ``taglen'' bytes of ``tag'' provided.  By default
``res'' is set to zero.  If all ``taglen'' bytes of ``tag'' can be verified then ``res'' is set to one (authenticated
message).

To make life simpler the following two functions are provided for memory bound OCB.

\index{ocb\_encrypt\_authenticate\_memory()}
\begin{verbatim}
int ocb_encrypt_authenticate_memory(int cipher,
    const unsigned char *key,    unsigned long keylen,
    const unsigned char *nonce,  
    const unsigned char *pt,     unsigned long ptlen,
          unsigned char *ct,
          unsigned char *tag,    unsigned long *taglen);
\end{verbatim}

This will OCB encrypt the message ``pt'' of length ``ptlen'' and store the ciphertext in ``ct''.  The length ``ptlen''
can be any arbitrary length.  

\index{ocb\_decrypt\_verify\_memory()}
\begin{verbatim}
int ocb_decrypt_verify_memory(int cipher,
    const unsigned char *key,    unsigned long keylen,
    const unsigned char *nonce,  
    const unsigned char *ct,     unsigned long ctlen,
          unsigned char *pt,
    const unsigned char *tag,    unsigned long taglen,
          int           *res);
\end{verbatim}

Similarly this will OCB decrypt and compare the internally computed tag against the tag provided. ``res'' is set 
appropriately.

\chapter{One-Way Cryptographic Hash Functions}
\section{Core Functions}

Like the ciphers there are hash core functions and a universal data type to hold the hash state called ``hash\_state''.  
To initialize hash XXX (where XXX is the name) call:
\index{Hash Functions}
\begin{verbatim}
void XXX_init(hash_state *md);
\end{verbatim}

This simply sets up the hash to the default state governed by the specifications of the hash.  To add data to the 
message being hashed call:
\begin{verbatim}
int XXX_process(hash_state *md, const unsigned char *in, unsigned long len);
\end{verbatim}

Essentially all hash messages are virtually infinitely\footnote{Most hashes are limited to $2^{64}$ bits or 2,305,843,009,213,693,952 bytes.} long message which 
are buffered.  The data can be passed in any sized chunks as long as the order of the bytes are the same the message digest
(hash output) will be the same.  For example, this means that:
\begin{verbatim}
md5_process(&md, "hello ", 6);
md5_process(&md, "world", 5);
\end{verbatim}
Will produce the same message digest as the single call:
\index{Message Digest}
\begin{verbatim}
md5_process(&md, "hello world", 11);
\end{verbatim}

To finally get the message digest (the hash) call:
\begin{verbatim}
int XXX_done(hash_state *md, 
              unsigned char *out);
\end{verbatim}

This function will finish up the hash and store the result in the ``out'' array.  You must ensure that ``out'' is long
enough for the hash in question.  Often hashes are used to get keys for symmetric ciphers so the ``XXX\_done()'' functions
will wipe the ``md'' variable before returning automatically.

To test a hash function call:
\begin{verbatim}
int XXX_test(void);
\end{verbatim}

This will return {\bf CRYPTO\_OK} if the hash matches the test vectors, otherwise it returns an error code.  An
example snippet that hashes a message with md5 is given below.
\begin{small}
\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
    hash_state md;
    unsigned char *in = "hello world", out[16];

    /* setup the hash */
    md5_init(&md);

    /* add the message */
    md5_process(&md, in, strlen(in));

    /* get the hash in out[0..15] */
    md5_done(&md, out);

    return 0;
}
\end{verbatim}
\end{small}

\section{Hash Descriptors}
Like the set of ciphers the set of hashes have descriptors too.  They are stored in an array called ``hash\_descriptor'' and
are defined by:
\begin{verbatim}
struct _hash_descriptor {
    char *name;
    unsigned long hashsize;    /* digest output size in bytes  */
    unsigned long blocksize;   /* the block size the hash uses */
    void (*init)   (hash_state *);
    int  (*process)(hash_state *, const unsigned char *, unsigned long);
    int  (*done)   (hash_state *, unsigned char *);
    int  (*test)   (void);
};
\end{verbatim}

Similarly ``name'' is the name of the hash function in ASCII (all lowercase).  ``hashsize'' is the size of the digest output
in bytes.  The remaining fields are pointers to the functions that do the respective tasks.  There is a function to
search the array as well called ``int find\_hash(char *name)''.  It returns -1 if the hash is not found, otherwise the
position in the descriptor table of the hash.

You can use the table to indirectly call a hash function that is chosen at runtime.  For example:
\begin{small}
\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   unsigned char buffer[100], hash[MAXBLOCKSIZE];
   int idx, x;
   hash_state md;

   /* register hashes .... */
   if (register_hash(&md5_desc) == -1) {
      printf("Error registering MD5.\n");
      return -1;
   }

   /* register other hashes ... */

   /* prompt for name and strip newline */
   printf("Enter hash name: \n");
   fgets(buffer, sizeof(buffer), stdin);
   buffer[strlen(buffer) - 1] = 0;

   /* get hash index */
   idx = find_hash(buffer);
   if (idx == -1) {
      printf("Invalid hash name!\n");
      return -1;
   }

   /* hash input until blank line */
   hash_descriptor[idx].init(&md);
   while (fgets(buffer, sizeof(buffer), stdin) != NULL)
         hash_descriptor[idx].process(&md, buffer, strlen(buffer));
   hash_descriptor[idx].done(&md, hash);

   /* dump to screen */
   for (x = 0; x < hash_descriptor[idx].hashsize; x++)
       printf("%02x ", hash[x]);
   printf("\n");
   return 0;
}
\end{verbatim}
\end{small}

Note the usage of ``MAXBLOCKSIZE''.  In Libtomcrypt no symmetric block, key or hash digest is larger than MAXBLOCKSIZE in
length.  This provides a simple size you can set your automatic arrays to that will not get overrun.

There are three helper functions as well:
\index{hash\_memory()} \index{hash\_file()}
\begin{verbatim}
int hash_memory(int hash, const unsigned char *data, 
                unsigned long len, unsigned char *dst,
                unsigned long *outlen);

int hash_file(int hash, const char *fname, 
              unsigned char *dst,
              unsigned long *outlen);

int hash_filehandle(int hash, FILE *in, 
                    unsigned char *dst, unsigned long *outlen);
\end{verbatim}

The ``hash'' parameter is the location in the descriptor table of the hash (\textit{e.g. the return of find\_hash()}).  
The ``*outlen'' variable is used to keep track of the output size.  You
must set it to the size of your output buffer before calling the functions.  When they complete succesfully they store
the length of the message digest back in it.  The functions are otherwise straightforward.  The ``hash\_filehandle'' 
function assumes that ``in'' is an file handle opened in binary mode.  It will hash to the end of file and not reset
the file position when finished.

To perform the above hash with md5 the following code could be used:
\begin{small}
\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   int idx, err;
   unsigned long len;
   unsigned char out[MAXBLOCKSIZE];

   /* register the hash */
   if (register_hash(&md5_desc) == -1) {
      printf("Error registering MD5.\n");
      return -1;
   }

   /* get the index of the hash  */
   idx = find_hash("md5");

   /* call the hash */
   len = sizeof(out);
   if ((err = hash_memory(idx, "hello world", 11, out, &len)) != CRYPT_OK) {
      printf("Error hashing data: %s\n", error_to_string(err));
      return -1;
   }
   return 0;
}
\end{verbatim}
\end{small}

The following hashes are provided as of this release:
\index{Hash descriptor table}
\begin{center}
\begin{tabular}{|c|c|c|}
      \hline Name & Descriptor Name & Size of Message Digest (bytes) \\
      \hline WHIRLPOOL & whirlpool\_desc & 64 \\
      \hline SHA-512 & sha512\_desc & 64 \\
      \hline SHA-384 & sha384\_desc & 48 \\
      \hline SHA-256 & sha256\_desc & 32 \\
      \hline SHA-224 & sha224\_desc & 28 \\
      \hline TIGER-192 & tiger\_desc & 24 \\
      \hline SHA-1 & sha1\_desc & 20 \\
      \hline RIPEMD-160 & rmd160\_desc & 20 \\
      \hline RIPEMD-128 & rmd128\_desc & 16 \\
      \hline MD5 & md5\_desc & 16 \\
      \hline MD4 & md4\_desc & 16 \\
      \hline MD2 & md2\_desc & 16 \\
      \hline
\end{tabular}
\end{center}

Similar to the cipher descriptor table you must register your hash algorithms before you can use them.  These functions
work exactly like those of the cipher registration code.  The functions are:
\index{register\_hash()} \index{unregister\_hash()}
\begin{verbatim}
int register_hash(const struct _hash_descriptor *hash);
int unregister_hash(const struct _hash_descriptor *hash);
\end{verbatim}

\section{Cipher Hash Construction}
\index{Cipher Hash Construction}
An addition to the suite of hash functions is the ``Cipher Hash Construction'' or ``CHC'' mode.  In this mode
applicable block ciphers (such as AES) can be turned into hash functions that other LTC functions can use.  In 
particular this allows a cryptosystem to be designed using very few moving parts.

In order to use the CHC system the developer will have to take a few extra steps.  First the ``chc\_desc'' hash
descriptor must be registered with register\_hash().  At this point the CHC hash cannot be used to hash
data.  While it is in the hash system you still have to tell the CHC code which cipher to use.  This is accomplished
via the chc\_register() function.

\index{chc\_register()}
\begin{verbatim}
int chc_register(int cipher);
\end{verbatim}

A cipher has to be registered with CHC (and also in the cipher descriptor tables with 
register\_cipher()).  The chc\_register() function will bind a cipher to the CHC system.  Only one cipher can 
be bound to the CHC hash at a time.  There are additional requirements for the system to work.

\begin{enumerate}
   \item The cipher must have a block size greater than 64--bits.  
   \item The cipher must allow an input key the size of the block size.
\end{enumerate}

Example of using CHC with the AES block cipher.

\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   int err; 

   /* register cipher and hash */
   if (register_cipher(&aes_enc_desc) == -1) {
      printf("Could not register cipher\n");
      return EXIT_FAILURE;
   }
   if (register_hash(&chc_desc) == -1) {
      printf("Could not register hash\n");
      return EXIT_FAILURE;
   }

   /* start chc with AES */
   if ((err = chc_register(find_cipher("aes"))) != CRYPT_OK) {
      printf("Error binding AES to CHC: %s\n", error_to_string(err));
   }

   /* now you can use chc_hash in any LTC function [aside from pkcs...] */
   /* ... */
\end{verbatim}


\section{Notice}
It is highly recommended that you \textbf{not} use the MD4 or MD5 hashes for the purposes of digital signatures or authentication codes.  
These hashes are provided for completeness and they still can be used for the purposes of password hashing or one-way accumulators
(e.g. Yarrow).

The other hashes such as the SHA-1, SHA-2 (that includes SHA-512, SHA-384 and SHA-256) and TIGER-192 are still considered secure
for all purposes you would normally use a hash for.

\chapter{Message Authentication Codes}
\section{HMAC Protocol}
Thanks to Dobes Vandermeer the library now includes support for hash based message authenication codes or HMAC for short.  An HMAC
of a message is a keyed authenication code that only the owner of a private symmetric key will be able to verify.  The purpose is
to allow an owner of a private symmetric key to produce an HMAC on a message then later verify if it is correct.  Any impostor or
eavesdropper will not be able to verify the authenticity of a message.  

The HMAC support works much like the normal hash functions except that the initialization routine requires you to pass a key 
and its length.  The key is much like a key you would pass to a cipher.  That is, it is simply an array of octets stored in
chars.  The initialization routine is:
\index{hmac\_init()}
\begin{verbatim}
int hmac_init(hmac_state *hmac, int hash, 
              const unsigned char *key, unsigned long keylen);
\end{verbatim}
The ``hmac'' parameter is the state for the HMAC code.  ``hash'' is the index into the descriptor table of the hash you want
to use to authenticate the message.  ``key'' is the pointer to the array of chars that make up the key.  ``keylen'' is the
length (in octets) of the key you want to use to authenticate the message.  To send octets of a message through the HMAC system you must use the following function:
\index{hmac\_process()}
\begin{verbatim}
int hmac_process(hmac_state *hmac, const unsigned char *buf,
                  unsigned long len);
\end{verbatim}
``hmac'' is the HMAC state you are working with. ``buf'' is the array of octets to send into the HMAC process.  ``len'' is the
number of octets to process.  Like the hash process routines you can send the data in arbitrarly sized chunks. When you 
are finished with the HMAC process you must call the following function to get the HMAC code:
\index{hmac\_done()}
\begin{verbatim}
int hmac_done(hmac_state *hmac, unsigned char *hashOut,
              unsigned long *outlen);
\end{verbatim}
``hmac'' is the HMAC state you are working with.  ``hashOut'' is the array of octets where the HMAC code should be stored.  You must
set ``outlen'' to the size of the destination buffer before calling this function.  It is updated with the length of the HMAC code
produced (depending on which hash was picked).  If ``outlen'' is less than the size of the message digest (and ultimately
the HMAC code) then the HMAC code is truncated as per FIPS-198 specifications (e.g. take the first ``outlen'' bytes).

There are two  utility functions provided to make using HMACs easier todo.  They accept the key and information about the
message (file pointer, address in memory) and produce the HMAC result in one shot.  These are useful if you want to avoid
calling the three step process yourself.

\index{hmac\_memory()}
\begin{verbatim}
int hmac_memory(int hash, const unsigned char *key, unsigned long keylen,
                const unsigned char *data, unsigned long len, 
                unsigned char *dst, unsigned long *dstlen);
\end{verbatim}
This will produce an HMAC code for the array of octets in ``data'' of length ``len''.  The index into the hash descriptor 
table must be provided in ``hash''.  It uses the key from ``key'' with a key length of ``keylen''.  
The result is stored in the array of octets ``dst'' and the length in ``dstlen''.  The value of ``dstlen'' must be set
to the size of the destination buffer before calling this function.  Similarly for files there is the  following function:
\index{hmac\_file()}
\begin{verbatim}
int hmac_file(int hash, const char *fname, const unsigned char *key,
              unsigned long keylen, 
              unsigned char *dst, unsigned long *dstlen);
\end{verbatim}
``hash'' is the index into the hash descriptor table of the hash you want to use.  ``fname'' is the filename to process.  
``key'' is the array of octets to use as the key of length ``keylen''.  ``dst'' is the array of octets where the 
result should be stored.

To test if the HMAC code is working there is the following function:
\index{hmac\_test()}
\begin{verbatim}
int hmac_test(void);
\end{verbatim}
Which returns {\bf CRYPT\_OK} if the code passes otherwise it returns an error code.  Some example code for using the 
HMAC system is given below.

\begin{small}
\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   int idx, err;
   hmac_state hmac;
   unsigned char key[16], dst[MAXBLOCKSIZE];
   unsigned long dstlen;

   /* register SHA-1 */
   if (register_hash(&sha1_desc) == -1) {
      printf("Error registering SHA1\n");
      return -1;
   }

   /* get index of SHA1 in hash descriptor table */
   idx = find_hash("sha1");

   /* we would make up our symmetric key in "key[]" here */

   /* start the HMAC */
   if ((err = hmac_init(&hmac, idx, key, 16)) != CRYPT_OK) {
      printf("Error setting up hmac: %s\n", error_to_string(err));
      return -1;
   }

   /* process a few octets */
   if((err = hmac_process(&hmac, "hello", 5) != CRYPT_OK) {
      printf("Error processing hmac: %s\n", error_to_string(err));
      return -1;
   }

   /* get result (presumably to use it somehow...) */
   dstlen = sizeof(dst);
   if ((err = hmac_done(&hmac, dst, &dstlen)) != CRYPT_OK) {
      printf("Error finishing hmac: %s\n", error_to_string(err));
      return -1;
   }
   printf("The hmac is %lu bytes long\n", dstlen);
  
   /* return */
   return 0;
}
\end{verbatim}
\end{small}

\section{OMAC Support}
OMAC\footnote{\url{http://crypt.cis.ibaraki.ac.jp/omac/omac.html}}, which stands for \textit{One-Key CBC MAC} is an 
algorithm which produces a Message Authentication Code (MAC) using only a block cipher such as AES.  From an API 
standpoint the OMAC routines work much like the HMAC routines do.  Instead in this case a cipher is used instead of a hash.  

To start an OMAC state you call
\index{omac\_init()}
\begin{verbatim}
int omac_init(omac_state *omac, int cipher, 
              const unsigned char *key, unsigned long keylen);
\end{verbatim}
The ``omac'' variable is the state for the OMAC algorithm.  ``cipher'' is the index into the cipher\_descriptor table
of the cipher\footnote{The cipher must have a 64 or 128 bit block size.  Such as CAST5, Blowfish, DES, AES, Twofish, etc.} you
wish to use.  ``key'' and ``keylen'' are the keys used to authenticate the data.

To send data through the algorithm call
\index{omac\_process()}
\begin{verbatim}
int omac_process(omac_state *state, 
                 const unsigned char *buf, unsigned long len);
\end{verbatim}
This will send ``len'' bytes from ``buf'' through the active OMAC state ``state''.  Returns \textbf{CRYPT\_OK} if the 
function succeeds.  The function is not sensitive to the granularity of the data.  For example,

\begin{verbatim}
omac_process(&mystate, "hello",  5);
omac_process(&mystate, " world", 6);
\end{verbatim}

Would produce the same result as,

\begin{verbatim}
omac_process(&mystate, "hello world",  11);
\end{verbatim}

When you are done processing the message you can call the following to compute the message tag.

\index{omac\_done()}
\begin{verbatim}
int omac_done(omac_state *state, 
              unsigned char *out, unsigned long *outlen);
\end{verbatim}
Which will terminate the OMAC and output the \textit{tag} (MAC) to ``out''.  Note that unlike the HMAC and other code 
``outlen'' can be smaller than the default MAC size (for instance AES would make a 16-byte tag).  Part of the OMAC 
specification states that the output may be truncated.  So if you pass in $outlen = 5$ and use AES as your cipher than
the output MAC code will only be five bytes long.  If ``outlen'' is larger than the default size it is set to the default
size to show how many bytes were actually used.

Similar to the HMAC code the file and memory functions are also provided.  To OMAC a buffer of memory in one shot use the 
following function.

\index{omac\_memory()}
\begin{verbatim}
int omac_memory(int cipher, 
                const unsigned char *key, unsigned long keylen,
                const unsigned char *msg, unsigned long msglen,
                unsigned char *out, unsigned long *outlen);
\end{verbatim}
This will compute the OMAC of ``msglen'' bytes of ``msg'' using the key ``key'' of length ``keylen'' bytes and the cipher
specified by the ``cipher'''th entry in the cipher\_descriptor table.  It will store the MAC in ``out'' with the same
rules as omac\_done.

To OMAC a file use
\index{omac\_file()}
\begin{verbatim}
int omac_file(int cipher, 
              const unsigned char *key, unsigned long keylen,
              const char *filename, 
              unsigned char *out, unsigned long *outlen);
\end{verbatim}

Which will OMAC the entire contents of the file specified by ``filename'' using the key ``key'' of length ``keylen'' bytes
and the cipher specified by the ``cipher'''th entry in the cipher\_descriptor table.  It will store the MAC in ``out'' with 
the same rules as omac\_done.

To test if the OMAC code is working there is the following function:
\index{omac\_test()}
\begin{verbatim}
int omac_test(void);
\end{verbatim}
Which returns {\bf CRYPT\_OK} if the code passes otherwise it returns an error code.  Some example code for using the 
OMAC system is given below.

\begin{small}
\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   int idx, err;
   omac_state omac;
   unsigned char key[16], dst[MAXBLOCKSIZE];
   unsigned long dstlen;

   /* register Rijndael */
   if (register_cipher(&rijndael_desc) == -1) {
      printf("Error registering Rijndael\n");
      return -1;
   }

   /* get index of Rijndael in cipher descriptor table */
   idx = find_cipher("rijndael");

   /* we would make up our symmetric key in "key[]" here */

   /* start the OMAC */
   if ((err = omac_init(&omac, idx, key, 16)) != CRYPT_OK) {
      printf("Error setting up omac: %s\n", error_to_string(err));
      return -1;
   }

   /* process a few octets */
   if((err = omac_process(&omac, "hello", 5) != CRYPT_OK) {
      printf("Error processing omac: %s\n", error_to_string(err));
      return -1;
   }

   /* get result (presumably to use it somehow...) */
   dstlen = sizeof(dst);
   if ((err = omac_done(&omac, dst, &dstlen)) != CRYPT_OK) {
      printf("Error finishing omac: %s\n", error_to_string(err));
      return -1;
   }
   printf("The omac is %lu bytes long\n", dstlen);
  
   /* return */
   return 0;
}
\end{verbatim}
\end{small}

\section{PMAC Support}
The PMAC\footnote{J.Black, P.Rogaway, ``A Block--Cipher Mode of Operation for Parallelizable Message Authentication''} 
protocol is another MAC algorithm that relies solely on a symmetric-key block cipher.  It uses essentially the same
API as the provided OMAC code.  

A PMAC state is initialized with the following.

\index{pmac\_init()}
\begin{verbatim}
int pmac_init(pmac_state *pmac, int cipher, 
              const unsigned char *key, unsigned long keylen);
\end{verbatim}
Which initializes the ``pmac'' state with the given ``cipher'' and ``key'' of length ``keylen'' bytes.  The chosen cipher
must have a 64 or 128 bit block size (e.x. AES).

To MAC data simply send it through the process function.

\index{pmac\_process()}
\begin{verbatim}
int pmac_process(pmac_state *state, 
                 const unsigned char *buf, unsigned long len);
\end{verbatim}
This will process ``len'' bytes of ``buf'' in the given ``state''.  The function is not sensitive to the granularity of the
data.  For example,

\begin{verbatim}
pmac_process(&mystate, "hello",  5);
pmac_process(&mystate, " world", 6);
\end{verbatim}

Would produce the same result as,

\begin{verbatim}
pmac_process(&mystate, "hello world",  11);
\end{verbatim}

When a complete message has been processed the following function can be called to compute the message tag.

\index{pmac\_done()}
\begin{verbatim}
int pmac_done(pmac_state *state, 
              unsigned char *out, unsigned long *outlen);
\end{verbatim}
This will store upto ``outlen'' bytes of the tag for the given ``state'' into ``out''.  Note that if ``outlen'' is larger
than the size of the tag it is set to the amount of bytes stored in ``out''.

Similar to the PMAC code the file and memory functions are also provided.  To PMAC a buffer of memory in one shot use the 
following function.

\index{pmac\_memory()}
\begin{verbatim}
int pmac_memory(int cipher, 
                const unsigned char *key, unsigned long keylen,
                const unsigned char *msg, unsigned long msglen,
                unsigned char *out, unsigned long *outlen);
\end{verbatim}
This will compute the PMAC of ``msglen'' bytes of ``msg'' using the key ``key'' of length ``keylen'' bytes and the cipher
specified by the ``cipher'''th entry in the cipher\_descriptor table.  It will store the MAC in ``out'' with the same
rules as omac\_done.

To PMAC a file use
\index{pmac\_file()}
\begin{verbatim}
int pmac_file(int cipher, 
              const unsigned char *key, unsigned long keylen,
              const char *filename, 
              unsigned char *out, unsigned long *outlen);
\end{verbatim}

Which will PMAC the entire contents of the file specified by ``filename'' using the key ``key'' of length ``keylen'' bytes
and the cipher specified by the ``cipher'''th entry in the cipher\_descriptor table.  It will store the MAC in ``out'' with 
the same rules as omac\_done.

To test if the PMAC code is working there is the following function:
\begin{verbatim}
int pmac_test(void);
\end{verbatim}
Which returns {\bf CRYPT\_OK} if the code passes otherwise it returns an error code.





\chapter{Pseudo-Random Number Generators}
\section{Core Functions}
The library provides an array of core functions for Pseudo-Random Number Generators (PRNGs) as well.  A cryptographic PRNG is
used to expand a shorter bit string into a longer bit string.  PRNGs are used wherever random data is required such as Public Key (PK)
key generation.  There is a universal structure called ``prng\_state''.  To initialize a PRNG call:
\index{PRNG start}
\begin{verbatim}
int XXX_start(prng_state *prng);
\end{verbatim}

This will setup the PRNG for future use and not seed it.  In order 
for the PRNG to be cryptographically useful you must give it entropy.  Ideally you'd have some OS level source to tap 
like in UNIX (see section 5.3).  To add entropy to the PRNG call:
\index{PRNG add\_entropy}
\begin{verbatim}
int XXX_add_entropy(const unsigned char *in, unsigned long len, 
                    prng_state *prng);
\end{verbatim}

Which returns {\bf CRYPTO\_OK} if the entropy was accepted.  Once you think you have enough entropy you call another
function to put the entropy into action.
\index{PRNG ready}
\begin{verbatim}
int XXX_ready(prng_state *prng);
\end{verbatim}

Which returns {\bf CRYPTO\_OK} if it is ready.  Finally to actually read bytes call:
\index{PRNG read}
\begin{verbatim}
unsigned long XXX_read(unsigned char *out, unsigned long len,
                       prng_state *prng);
\end{verbatim}

Which returns the number of bytes read from the PRNG.  When you are finished with a PRNG state you call
the following.

\index{PRNG done}
\begin{verbatim}
void XXX_done(prng_state *prng);
\end{verbatim}

This will terminate a PRNG state and free any memory (if any) allocated.  To export a PRNG state
so that you can later resume the PRNG call the following.

\index{PRNG export}
\begin{verbatim}
int XXX_export(unsigned char *out, unsigned long *outlen, 
               prng_state    *prng);
\end{verbatim}

This will write a ``PRNG state'' to the buffer ``out'' of length ``outlen'' bytes.  The idea of 
the export is meant to be used as a ``seed file''.  That is, when the program starts up there will not likely
be that much entropy available.   To import a state to seed a PRNG call the following function.

\index{PRNG import}
\begin{verbatim}
int XXX_import(const unsigned char *in, unsigned long inlen, 
                     prng_state     *prng);
\end{verbatim}

This will call the start and add\_entropy functions of the given PRNG.  It will use the state in
``in'' of length ``inlen'' as the initial seed.  You must pass the same seed length as was exported
by the corresponding export function.

Note that importing a state will not ``resume'' the PRNG from where it left off.  That is, if you export
a state, emit (say) 8 bytes and then import the previously exported state the next 8 bytes will not 
specifically equal the 8 bytes you generated previously.

When a program is first executed the normal course of operation is 

\begin{enumerate}
   \item Gather entropy from your sources for a given period of time or number of events.
   \item Start, use your entropy via add\_entropy and ready the PRNG yourself.
\end{enumerate}

When your program is finished you simply call the export function and save the state to a medium (disk,
flash memory, etc).  The next time your application starts up you can detect the state, feed it to the 
import function and go on your way.  It is ideal that (as soon as possible) after startup you export a
fresh state.  This helps in the case that the program aborts or the machine is powered down without
being given a chance to exit properly.  

Note that even if you have a state to import it is important to add new entropy to the state.  However,
there is less pressure to do so.  

To test a PRNG for operational conformity call the following functions.

\index{PRNG test}
\begin{verbatim}
int XXX_test(void);
\end{verbatim}

This will return \textbf{CRYPT\_OK} if PRNG is operating properly.

\subsection{Remarks}

It is possible to be adding entropy and reading from a PRNG at the same time.  For example, if you first seed the PRNG
and call ready() you can now read from it.  You can also keep adding new entropy to it.  The new entropy will not be used
in the PRNG until ready() is called again.  This allows the PRNG to be used and re-seeded at the same time.  No real error 
checking is guaranteed to see if the entropy is sufficient or if the PRNG is even in a ready state before reading.

\subsection{Example}

Below is a simple snippet to read 10 bytes from yarrow.  Its important to note that this snippet is 
{\bf NOT} secure since the entropy added is not random.

\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   prng_state prng;
   unsigned char buf[10];
   int err;
   
   /* start it */
   if ((err = yarrow_start(&prng)) != CRYPT_OK) {
      printf("Start error: %s\n", error_to_string(err));
   }
   /* add entropy */
   if ((err = yarrow_add_entropy("hello world", 11, &prng)) != CRYPT_OK) {
      printf("Add_entropy error: %s\n", error_to_string(err));
   }
   /* ready and read */
   if ((err = yarrow_ready(&prng)) != CRYPT_OK) {
      printf("Ready error: %s\n", error_to_string(err));
   }
   printf("Read %lu bytes from yarrow\n", yarrow_read(buf, 10, &prng));
   return 0;
}
\end{verbatim}

\section{PRNG Descriptors}
\index{PRNG Descriptor}
PRNGs have descriptors too (surprised?). Stored in the structure ``prng\_descriptor''.  The format of an element is:
\begin{verbatim}
struct _prng_descriptor {
    char *name;
    int  export_size;    /* size in bytes of exported state */
    int (*start)      (prng_state *);
    int (*add_entropy)(const unsigned char *, unsigned long, prng_state *);
    int (*ready)      (prng_state *);
    unsigned long (*read)(unsigned char *, unsigned long len, prng_state *);
    void (*done)(prng_state *);
    int (*export)(unsigned char *, unsigned long *, prng_state *);
    int (*import)(const unsigned char *, unsigned long, prng_state *);
    int (*test)(void);
};
\end{verbatim}

There is a ``int find\_prng(char *name)'' function as well.  Returns -1 if the PRNG is not found, otherwise it returns
the position in the prng\_descriptor array.

Just like the ciphers and hashes you must register your prng before you can use it.  The two functions provided work
exactly as those for the cipher registry functions.  They are:
\begin{verbatim}
int register_prng(const struct _prng_descriptor *prng);
int unregister_prng(const struct _prng_descriptor *prng);
\end{verbatim}

\subsection{PRNGs Provided}
\begin{figure}[here]
\begin{center}
\begin{small}
\begin{tabular}{|c|c|l|}
\hline \textbf{Name} & \textbf{Descriptor} & \textbf{Usage} \\
\hline Yarrow & yarrow\_desc & Fast short-term PRNG \\
\hline Fortuna & fortuna\_desc & Fast long-term PRNG (recommended) \\
\hline RC4 & rc4\_desc & Stream Cipher \\
\hline SOBER-128 & sober128\_desc & Stream Cipher (also very fast PRNG) \\
\hline
\end{tabular}
\end{small}
\end{center}
\caption{List of Provided PRNGs}
\end{figure}

\subsubsection{Yarrow}
Yarrow is fast PRNG meant to collect an unspecified amount of entropy from sources 
(keyboard, mouse, interrupts, etc) and produce an unbounded string of random bytes.  

\textit{Note:} This PRNG is still secure for most taskings but is no longer recommended.  Users
should use Fortuna instead.

\subsubsection{Fortuna}

Fortuna is a fast attack tolerant and more thoroughly designed PRNG suitable for long term
usage.  It is faster than the default implementation of Yarrow\footnote{Yarrow has been implemented
to work with most cipher and hash combos based on which you have chosen to build into the library.} while
providing more security.  

Fortuna is slightly less flexible than Yarrow in the sense that it only works with the AES block cipher 
and SHA--256 hash function.  Technically Fortuna will work with any block cipher that accepts a 256--bit
key and any hash that produces at least a 256--bit output.  However, to make the implementation simpler
it has been fixed to those choices.

Fortuna is more secure than Yarrow in the sense that attackers who learn parts of the entropy being 
added to the PRNG learn far less about the state than that of Yarrow.  Without getting into to many
details Fortuna has the ability to recover from state determination attacks where the attacker starts
to learn information from the PRNGs output about the internal state.  Yarrow on the other hand cannot 
recover from that problem until new entropy is added to the pool and put to use through the ready() function.

\subsubsection{RC4}

RC4 is an old stream cipher that can also double duty as a PRNG in a pinch.  You ``key'' it by
calling add\_entropy() and setup the key by calling ready().  You can only add upto 256 bytes via
add\_entropy().  

When you read from RC4 the output of the RC4 algorithm is XOR'd against your buffer you provide.  In this
manner you can use rc4\_read() as an encrypt (and decrypt) function.  

You really shouldn't use RC4 anymore.  This isn't because RC4 is weak (though biases are known to exist) just
simply that faster alternatives exist.

\subsubsection{SOBER-128}

SOBER-128 is a stream cipher designed by the QUALCOMM Australia team.  Like RC4 you ``key'' it by 
calling add\_entropy().  There is no need to call ready() for this PRNG as it does not do anything.  

Note that this cipher has several oddities about how it operates.  The first time you call 
add\_entropy() that sets the cipher's key.  Every other time you call the same function it sets
the cipher's IV variable.  The IV mechanism allows you to encrypt several messages with the same
key and not re--use the same key material.

Unlike Yarrow and Fortuna all of the entropy (and hence security) of this algorithm rests in the data
you pass it on the first call to add\_entropy().  All buffers sent to add\_entropy() must have a length
that is a multiple of four bytes.

Like RC4 the output of SOBER--128 is XOR'ed against the buffer you provide it.  In this manner you can use
sober128\_read() as an encrypt (and decrypt) function.

Since SOBER-128 has a fixed keying scheme and is very fast (faster than RC4) the ideal usage of SOBER-128 is to 
key it from the output of Fortuna (or Yarrow) and use it to encrypt messages.  It is also ideal for
simulations which need a high quality (and fast) stream of bytes.  

\subsubsection{Example Usage}
\begin{small}
\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   prng_state prng;
   unsigned char buf[32];
   int err;

   if ((err = rc4_start(&prng)) != CRYPT_OK) {
      printf("RC4 init error: %s\n", error_to_string(err));
      exit(-1);
   }

   /* use ``key'' as the key */
   if ((err = rc4_add_entropy("key", 3, &prng)) != CRYPT_OK) {
      printf("RC4 add entropy error: %s\n", error_to_string(err));
      exit(-1);
   }

   /* setup RC4 for use */
   if ((err = rc4_ready(&prng)) != CRYPT_OK) {
      printf("RC4 ready error: %s\n", error_to_string(err));
      exit(-1);
   }

   /* encrypt buffer */
   strcpy(buf,"hello world");
   if (rc4_read(buf, 11, &prng) != 11) {
      printf("RC4 read error\n");
      exit(-1);
   }
   return 0;
}   
\end{verbatim}
\end{small}
To decrypt you have to do the exact same steps.  

\section{The Secure RNG}
\index{Secure RNG}
An RNG is related to a PRNG except that it doesn't expand a smaller seed to get the data.  They generate their random bits
by performing some computation on fresh input bits.  Possibly the hardest thing to get correctly in a cryptosystem is the 
PRNG.  Computers are deterministic beasts that try hard not to stray from pre-determined paths.  That makes gathering 
entropy needed to seed the PRNG a hard task.  

There is one small function that may help on certain platforms:
\index{rng\_get\_bytes()}
\begin{verbatim}
unsigned long rng_get_bytes(unsigned char *buf, unsigned long len, 
                  void (*callback)(void));
\end{verbatim}

Which will try one of three methods of getting random data.  The first is to open the popular ``/dev/random'' device which 
on most *NIX platforms provides cryptographic random bits\footnote{This device is available in Windows through the Cygwin compiler suite.  It emulates ``/dev/random'' via the Microsoft CSP.}.  
The second method is to try the Microsoft Cryptographic Service Provider and read the RNG.  The third method is an ANSI C 
clock drift method that is also somewhat popular but gives bits of lower entropy.  The ``callback'' parameter is a pointer to a function that returns void.  Its used when the slower ANSI C RNG must be 
used so the calling application can still work.  This is useful since the ANSI C RNG has a throughput of three 
bytes a second.  The callback pointer may be set to {\bf NULL} to avoid using it if you don't want to.  The function 
returns the number of bytes actually read from any RNG source.  There is a function to help setup a PRNG as well:
\index{rng\_make\_prng()}
\begin{verbatim}
int rng_make_prng(int bits, int wprng, prng_state *prng, 
                  void (*callback)(void));
\end{verbatim}
This will try to setup the prng with a state of at least ``bits'' of entropy.  The ``callback'' parameter works much like
the callback in ``rng\_get\_bytes()''.  It is highly recommended that you use this function to setup your PRNGs unless you have a
platform where the RNG doesn't work well.  Example usage of this function is given below.

\begin{small}
\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   ecc_key mykey;
   prng_state prng;
   int err;

   /* register yarrow */
   if (register_prng(&yarrow_desc) == -1) {
      printf("Error registering Yarrow\n");
      return -1;
   }

   /* setup the PRNG */
   if ((err = rng_make_prng(128, find_prng("yarrow"), &prng, NULL)) != CRYPT_OK) {
      printf("Error setting up PRNG, %s\n", error_to_string(err));
      return -1;
   }

   /* make a 192-bit ECC key */
   if ((err = ecc_make_key(&prng, find_prng("yarrow"), 24, &mykey)) != CRYPT_OK) {
      printf("Error making key: %s\n", error_to_string(err));
      return -1;
   }
   return 0;
}
\end{verbatim}
\end{small}

\subsection{The Secure PRNG Interface}
It is possible to access the secure RNG through the PRNG interface and in turn use it within dependent functions such
as the PK API.  This simplifies the cryptosystem on platforms where the secure RNG is fast.  The secure PRNG never 
requires to be started, that is you need not call the start, add\_entropy or ready functions.  For example, consider
the previous example using this PRNG.

\begin{small}
\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   ecc_key mykey;
   int err;

   /* register SPRNG */
   if (register_prng(&sprng_desc) == -1) {
      printf("Error registering SPRNG\n");
      return -1;
   }

   /* make a 192-bit ECC key */
   if ((err = ecc_make_key(NULL, find_prng("sprng"), 24, &mykey)) != CRYPT_OK) {
      printf("Error making key: %s\n", error_to_string(err));
      return -1;
   }
   return 0;
}
\end{verbatim}
\end{small}

\chapter{RSA Public Key Cryptography}

\section{Introduction}
RSA wrote the PKCS \#1 specifications which detail RSA Public Key Cryptography.  In the specifications are
padding algorithms for encryption and signatures.  The standard includes ``v1.5'' and ``v2.0'' algorithms.
To simplify matters a little the v2.0 encryption and signature padding algorithms are called OAEP and PSS 
respectively.  

\section{PKCS \#1 Encryption}

PKCS \#1 RSA Encryption amounts to OAEP padding of the input message followed by the modular exponentiation.  As far as this portion of
the library is concerned we are only dealing with th OAEP padding of the message.

\subsection{OAEP Encoding}

\index{pkcs\_1\_oaep\_encode()}
\begin{alltt}
int pkcs_1_oaep_encode(const unsigned char *msg,    unsigned long msglen,
                       const unsigned char *lparam, unsigned long lparamlen,
                             unsigned long modulus_bitlen, prng_state *prng,
                             int           prng_idx,         int  hash_idx,
                             unsigned char *out,    unsigned long *outlen);
\end{alltt}

This accepts ``msg'' as input of length ``msglen'' which will be OAEP padded.  The ``lparam'' variable is an additional system specific
tag that can be applied to the encoding.  This is useful to identify which system encoded the message.  If no variance is desired then
``lparam'' can be set to \textbf{NULL}.  

OAEP encoding requires the length of the modulus in bits in order to calculate the size of the output.  This is passed as the parameter
``modulus\_bitlen''.  ``hash\_idx'' is the index into the hash descriptor table of the hash desired.  PKCS \#1 allows any hash to be 
used but both the encoder and decoder must use the same hash in order for this to succeed.  The size of hash output affects the maximum
 sized input message.  ``prng\_idx'' and ``prng'' are the random number generator arguments required to randomize the padding process.  
The padded message is stored in ``out'' along with the length in ``outlen''.

If $h$ is the length of the hash and $m$ the length of the modulus (both in octets) then the maximum payload for ``msg'' is 
$m - 2h - 2$.  For example, with a $1024$--bit RSA key and SHA--1 as the hash the maximum payload is $86$ bytes.  

Note that when the message is padded it still has not been RSA encrypted.  You must pass the output of this function to 
rsa\_exptmod() to encrypt it. 

\subsection{OAEP Decoding}

\index{pkcs\_1\_oaep\_decode()}
\begin{alltt}
int pkcs_1_oaep_decode(const unsigned char *msg,    unsigned long msglen,
                       const unsigned char *lparam, unsigned long lparamlen,
                             unsigned long modulus_bitlen, int hash_idx,
                             unsigned char *out,    unsigned long *outlen,
                             int           *res);
\end{alltt}

This function decodes an OAEP encoded message and outputs the original message that was passed to the OAEP encoder.  ``msg'' is the 
output of pkcs\_1\_oaep\_encode() of length ``msglen''.  ``lparam'' is the same system variable passed to the OAEP encoder.  If it does not
match what was used during encoding this function will not decode the packet.  ``modulus\_bitlen'' is the size of the RSA modulus in bits
and must match what was used during encoding.  Similarly the ``hash\_idx'' index into the hash descriptor table must match what was used
during encoding.

If the function succeeds it decodes the OAEP encoded message into ``out'' of length ``outlen'' and stores a 
$1$ in ``res''.  If the packet is invalid it stores $0$ in ``res'' and if the function fails for another reason
it returns an error code.  

\subsection{PKCS \#1 v1.5 Encoding}

\index{pkcs\_1\_v15\_es\_encode()}
\begin{verbatim}
int pkcs_1_v15_es_encode(const unsigned char *msg,    unsigned long msglen,
                               unsigned long  modulus_bitlen, 
                               prng_state    *prng,   int           prng_idx,
                               unsigned char *out,    unsigned long *outlen);
\end{verbatim}

This will PKCS v1.5 encode the data in ``msg'' of length ``msglen''.  Pass the length (in bits) of your
RSA modulus in ``modulus\_bitlen''.  The encoded data will be stored in ``out'' of length ``outlen''.

\subsection{PKCS \#1 v1.5 Decoding}
\index{pkcs\_1\_v15\_es\_decode()}
\begin{verbatim}
int pkcs_1_v15_es_decode(const unsigned char *msg,  unsigned long msglen,
                               unsigned long modulus_bitlen,
                               unsigned char *out,  unsigned long outlen,
                               int           *res);
\end{verbatim}

This will PKCS v1.5 decode the message in ``msg'' of length ``msglen''.  It will store the output in ``out''. Note
that the length of the output ``outlen'' is a constant.  This decoder cannot determine the original message 
length.  If the data in ``msg'' is a valid packet then a $1$ is stored in ``res'', otherwise a $0$ is 
stored.

\section{PKCS \#1 Digital Signatures}

\subsection{PSS Encoding}
PSS encoding is the second half of the PKCS \#1 standard which is padding to be applied to messages that are signed.  

\index{pkcs\_1\_pss\_encode()}
\begin{alltt}
int pkcs_1_pss_encode(const unsigned char *msghash, unsigned long msghashlen,
                            unsigned long saltlen,  prng_state   *prng,     
                            int           prng_idx, int           hash_idx,
                            unsigned long modulus_bitlen,
                            unsigned char *out,     unsigned long *outlen);
\end{alltt}

This function assumes the message to be PSS encoded has previously been hashed.  The input hash ``msghash'' is of length 
``msghashlen''.  PSS allows a variable length random salt (it can be zero length) to be introduced in the signature process.  
``hash\_idx'' is the index into the hash descriptor table of the hash to use.  ``prng\_idx'' and ``prng'' are the random
number generator information required for the salt.

Similar to OAEP encoding ``modulus\_bitlen'' is the size of the RSA modulus (in bits).  It limits the size of the salt.  If $m$ is the length
of the modulus $h$ the length of the hash output (in octets) then there can be $m - h - 2$ bytes of salt.  

This function does not actually sign the data it merely pads the hash of a message so that it can be processed by rsa\_exptmod().

\subsection{PSS Decoding}

To decode a PSS encoded signature block you have to use the following.

\index{pkcs\_1\_pss\_decode()}
\begin{alltt}
int pkcs_1_pss_decode(const unsigned char *msghash, unsigned long msghashlen,
                      const unsigned char *sig,     unsigned long siglen,
                            unsigned long saltlen,  int           hash_idx,
                            unsigned long modulus_bitlen, int    *res);
\end{alltt}
This will decode the PSS encoded message in ``sig'' of length ``siglen'' and compare it to values in ``msghash'' of length
``msghashlen''.  If the block is a valid PSS block and the decoded hash equals the hash supplied ``res'' is set to non--zero.  Otherwise, 
it is set to zero.  The rest of the parameters are as in the PSS encode call.

It's important to use the same ``saltlen'' and hash for both encoding and decoding as otherwise the procedure will not work.

\subsection{PKCS \#1 v1.5 Encoding}

\index{pkcs\_1\_v15\_sa\_encode()}
\begin{verbatim}
int pkcs_1_v15_sa_encode(const unsigned char *msghash,  unsigned long msghashlen,
                               int            hash_idx, unsigned long modulus_bitlen,
                               unsigned char *out,      unsigned long *outlen);
\end{verbatim}

This will PKCS \#1 v1.5 signature encode the message hash ``msghash''  of length ``msghashlen''.  You have
to tell this routine which hash produced the message hash in ``hash\_idx''.  The encoded hash is stored
in ``out'' of length ``outlen''.

\subsection{PKCS \#1 v1.5 Decoding}

\index{pkcs\_1\_v15\_sa\_decode()}
\begin{verbatim}
int pkcs_1_v15_sa_decode(const unsigned char *msghash, unsigned long msghashlen,
                         const unsigned char *sig,     unsigned long siglen,
                               int           hash_idx, unsigned long modulus_bitlen, 
                               int          *res);
\end{verbatim}

This will PKCS \#1 v1.5 signature decode the data in ``sig'' of length ``siglen'' and compare the extracted
hash against ``msghash'' of length ``msghashlen''.  You have to tell this routine which hash produced the
message digest in ``hash\_idx''.  If the packet is valid and the hashes match ``res'' is set to $1$.  Otherwise,
it is set to $0$.

\section{RSA Operations}
\subsection{Background}

RSA is a public key algorithm that is based on the inability to find the ``e-th'' root modulo a composite of unknown 
factorization.  Normally the difficulty of breaking RSA is associated with the integer factoring problem but they are
not strictly equivalent.

The system begins with with two primes $p$ and $q$ and their product $N = pq$.  The order or ``Euler totient'' of the
multiplicative sub-group formed modulo $N$ is given as $\phi(N) = (p - 1)(q - 1)$ which can be reduced to 
$\mbox{lcm}(p - 1, q - 1)$.  The public key consists of the composite $N$ and some integer $e$ such that 
$\mbox{gcd}(e, \phi(N)) = 1$.  The private key consists of the composite $N$ and the inverse of $e$ modulo $\phi(N)$ 
often simply denoted as $de \equiv 1\mbox{ }(\mbox{mod }\phi(N))$.

A person who wants to encrypt with your public key simply forms an integer (the plaintext) $M$ such that 
$1 < M < N-2$ and computes the ciphertext $C = M^e\mbox{ }(\mbox{mod }N)$.  Since finding the inverse exponent $d$
given only $N$ and $e$ appears to be intractable only the owner of the private key can decrypt the ciphertext and compute
$C^d \equiv \left (M^e \right)^d \equiv M^1 \equiv M\mbox{ }(\mbox{mod }N)$.  Similarly the owner of the private key 
can sign a message by ``decrypting'' it.  Others can verify it by ``encrypting'' it.  

Currently RSA is a difficult system to cryptanalyze provided that both primes are large and not close to each other.  
Ideally $e$ should be larger than $100$ to prevent direct analysis.  For example, if $e$ is three and you do not pad
the plaintext to be encrypted than it is possible that $M^3 < N$ in which case finding the cube-root would be trivial.  
The most often suggested value for $e$ is $65537$ since it is large enough to make such attacks impossible and also well 
designed for fast exponentiation (requires 16 squarings and one multiplication).

It is important to pad the input to RSA since it has particular mathematical structure.  For instance  
$M_1^dM_2^d = (M_1M_2)^d$ which can be used to forge a signature.  Suppose $M_3 = M_1M_2$ is a message you want
to have a forged signature for.  Simply get the signatures for $M_1$ and $M_2$ on their own and multiply the result
together.  Similar tricks can be used to deduce plaintexts from ciphertexts.  It is important not only to sign 
the hash of documents only but also to pad the inputs with data to remove such structure.  

\subsection{RSA Key Generation}

For RSA routines a single ``rsa\_key'' structure is used.  To make a new RSA key call:
\index{rsa\_make\_key()}
\begin{verbatim}
int rsa_make_key(prng_state *prng, 
                 int wprng, int size, 
                 long e, rsa_key *key);
\end{verbatim}

Where ``wprng'' is the index into the PRNG descriptor array.  ``size'' is the size in bytes of the RSA modulus desired.
``e'' is the encryption exponent desired, typical values are 3, 17, 257 and 65537.  I suggest you stick with 65537 since its big
enough to prevent trivial math attacks and not super slow.  ``key'' is where the key is placed.  All keys must be at 
least 128 bytes and no more than 512 bytes in size (\textit{that is from 1024 to 4096 bits}).

Note that the ``rsa\_make\_key()'' function allocates memory at runtime when you make the key.  Make sure to call 
``rsa\_free()'' (see below) when you are finished with the key.  If ``rsa\_make\_key()'' fails it will automatically 
free the ram allocated itself.

\index{PK\_PRIVATE} \index{PK\_PUBLIC}
There are two types of RSA keys.  The types are {\bf PK\_PRIVATE} and {\bf PK\_PUBLIC}.  The first type is a private 
RSA key which includes the CRT parameters\footnote{As of v0.99 the PK\_PRIVATE\_OPTIMIZED type has been deprecated
and has been replaced by the PK\_PRIVATE type.} in the form of a RSAPrivateKey.  The second type is a public RSA key
which only includes the modulus and public exponent.  It takes the form of a RSAPublicKey.

\subsection{RSA Exponentiation}

To do raw work with the RSA function call:
\index{rsa\_exptmod()}
\begin{verbatim}
int rsa_exptmod(const unsigned char *in,   unsigned long inlen,
                      unsigned char *out,  unsigned long *outlen, int which,
                      prng_state    *prng, int           prng_idx,
                      rsa_key *key);
\end{verbatim}
This loads the bignum from ``in'' as a big endian word in the format PKCS specifies, raises it to either ``e'' or ``d'' and stores the result
in ``out'' and the size of the result in ``outlen''. ``which'' is set to {\bf PK\_PUBLIC} to use ``e'' 
(i.e. for encryption/verifying) and set to {\bf PK\_PRIVATE} to use ``d'' as the exponent (i.e. for decrypting/signing).

Note that the output of his function is zero-padded as per PKCS \#1 specifications.  This allows this routine to 
interoprate with PKCS \#1 padding functions properly.

\subsection{RSA Key Encryption}
Normally RSA is used to encrypt short symmetric keys which are then used in block ciphers to encrypt a message.
To facilitate encrypting short keys the following functions have been provided.

\index{rsa\_encrypt\_key()}
\begin{verbatim}
int rsa_encrypt_key(const unsigned char *inkey,  unsigned long inlen,
                          unsigned char *outkey, unsigned long *outlen,
                    const unsigned char *lparam, unsigned long lparamlen,
                    prng_state *prng, int prng_idx, int hash_idx, rsa_key *key);
\end{verbatim}
This function will OAEP pad ``inkey'' of length inlen bytes then RSA encrypt it and store the ciphertext
in ``outkey'' of length ``outlen''.  The ``lparam'' and ``lparamlen'' are the same parameters you would pass
to pkcs\_1\_oaep\_encode().

\index{rsa\_decrypt\_key()}
\begin{verbatim}
int rsa_decrypt_key(const unsigned char *in,     unsigned long inlen,
                          unsigned char *outkey, unsigned long *keylen, 
                    const unsigned char *lparam, unsigned long lparamlen,
                          prng_state    *prng,   int           prng_idx,
                          int            hash_idx, int *res,
                          rsa_key       *key);
\end{verbatim}
This function will RSA decrypt ``in'' of length ``inlen'' then OAEP depad the resulting data and store it in
``outkey'' of length ``outlen''.  The ``lparam'' and ``lparamlen'' are the same parameters you would pass
to pkcs\_1\_oaep\_decode().

If the RSA decrypted data isn't a valid OAEP packet then ``res'' is set to $0$.  Otherwise, it is set to $1$.

\subsection{RSA Hash Signatures}
Similar to RSA key encryption RSA is also used to ``digitally sign'' message digests (hashes).  To facilitate this
process the following functions have been provided.

\index{rsa\_sign\_hash()}
\begin{verbatim}
int rsa_sign_hash(const unsigned char *msghash,  unsigned long  msghashlen, 
                        unsigned char *sig,      unsigned long *siglen, 
                        prng_state    *prng,     int            prng_idx,
                        int            hash_idx, unsigned long  saltlen,
                        rsa_key *key);
\end{verbatim}

This will PSS encode the message hash ``msghash'' of length ``msghashlen''.  Next the PSS encoded message is
RSA ``signed'' and the output is stored in ``sig'' of length ``siglen''.  


\index{rsa\_verify\_hash()}
\begin{verbatim}
int rsa_verify_hash(const unsigned char *sig,      unsigned long siglen,
                    const unsigned char *msghash,  unsigned long msghashlen,
                          prng_state    *prng,     int           prng_idx,
                          int            hash_idx, unsigned long saltlen,
                          int           *stat,     rsa_key      *key);
\end{verbatim}

This will RSA ``verify'' the signature in ``sig'' of length ``siglen''.  Next the RSA decoded data is PSS decoded
and the extracted hash is compared against the message hash ``msghash'' of length ``msghashlen''.

If the RSA decoded data is not a valid PSS message or if the PSS decoded hash does not match the ``msghash'' 
the value ``res'' is set to $0$.  Otherwise, if the function succeeds and signature is valid ``res'' is set
to $1$.

\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   int           err, hash_idx, prng_idx, res;
   unsigned long l1, l2;
   unsigned char pt[16], pt2[16], out[1024];
   rsa_key       key;

   /* register prng/hash */
   if (register_prng(&sprng_desc) == -1) {
      printf("Error registering sprng");
      return EXIT_FAILURE;
   }

   if (register_hash(&sha1_desc) == -1) {
      printf("Error registering sha1");
      return EXIT_FAILURE;
   }
   hash_idx = find_hash("sha1");
   prng_idx = find_prng("sprng");

   /* make an RSA-1024 key */
   if ((err = rsa_make_key(NULL,     /* PRNG state */
                           prng_idx, /* PRNG idx */
                           1024/8,   /* 1024-bit key */
                           65537,    /* we like e=65537 */
                           &key)     /* where to store the key */
       ) != CRYPT_OK) {
       printf("rsa_make_key %s", error_to_string(err));
       return EXIT_FAILURE;
   }

   /* fill in pt[] with a key we want to send ... */
   l1 = sizeof(out);
   if ((err = rsa_encrypt_key(pt,    /* data we wish to encrypt */
                              16,    /* data is 16 bytes long */
                             out,    /* where to store ciphertext */
                             &l1,    /* length of ciphertext */
                       "TestApp",    /* our lparam for this program */
                               7,    /* lparam is 7 bytes long */
                            NULL,    /* PRNG state */
                        prng_idx,    /* prng idx */
                        hash_idx,    /* hash idx */
                            &key)    /* our RSA key */
       ) != CRYPT_OK) {
       printf("rsa_encrypt_key %s", error_to_string(err));
       return EXIT_FAILURE;
   }

   /* now let's decrypt the encrypted key */
   l2 = sizeof(pt2);
   if ((err = rsa_decrypt_key(out, /* encrypted data */
                               l1, /* length of ciphertext */
                              pt2, /* where to put plaintext */
                              &l2, /* plaintext length */
                        "TestApp", /* lparam for this program */
                                7, /* lparam is 7 bytes long */
                             NULL, /* PRNG state */
                         prng_idx, /* prng idx */
                         hash_idx, /* hash idx */
                             &res, /* validity of data */
                             &key) /* our RSA key */ 
        ) != CRYPT_OK) {
       printf("rsa_decrypt_key %s", error_to_string(err));
       return EXIT_FAILURE;
   }
   /* if all went well pt == pt2, l2 == 16, res == 1 */
}
\end{verbatim}


\chapter{Diffie-Hellman Key Exchange}

\section{Background}

Diffie-Hellman was the original public key system proposed.  The system is based upon the group structure
of finite fields.  For Diffie-Hellman a prime $p$ is chosen and a ``base'' $b$ such that $b^x\mbox{ }(\mbox{mod }p)$ 
generates a large sub-group of prime order (for unique values of $x$).

A secret key is an exponent $x$ and a public key is the value of $y \equiv g^x\mbox{ }(\mbox{mod }p)$.  The term
``discrete logarithm'' denotes the action of finding $x$ given only $y$, $g$ and $p$.  The key exchange part of
Diffie-Hellman arises from the fact that two users A and B with keys $(A_x, A_y)$ and $(B_x, B_y)$ can exchange 
a shared key $K \equiv B_y^{A_x} \equiv A_y^{B_x} \equiv g^{A_xB_x}\mbox{ }(\mbox{mod }p)$.

From this public encryption and signatures can be developed.  The trivial way to encrypt (for example) using a public key 
$y$ is to perform the key exchange offline.  The sender invents a key $k$ and its public copy 
$k' \equiv g^k\mbox{ }(\mbox{mod }p)$ and uses $K \equiv k'^{A_x}\mbox{ }(\mbox{mod }p)$ as a key to encrypt
the message with.  Typically $K$ would be sent to a one-way hash and the message digested used as a key in a 
symmetric cipher.

It is important that the order of the sub-group that $g$ generates not only be large but also prime.  There are
discrete logarithm algorithms that take $\sqrt r$ time given the order $r$.  The discrete logarithm can be computed
modulo each prime factor of $r$ and the results combined using the Chinese Remainder Theorem.  In the cases where 
$r$ is ``B-Smooth'' (e.g. all small factors or powers of small prime factors) the solution is trivial to find.

To thwart such attacks the primes and bases in the library have been designed and fixed.  Given a prime $p$ the order of
 the sub-group generated is a large prime namely ${p - 1} \over 2$.  Such primes are known as ``strong primes'' and the 
smaller prime (e.g. the order of the base) are known as Sophie-Germaine primes.

\section{Core Functions}

This library also provides core Diffie-Hellman functions so you can negotiate keys over insecure mediums.  The routines 
provided are relatively easy to use and only take two function calls to negotiate a shared key.  There is a structure
called ``dh\_key'' which stores the Diffie-Hellman key in a format these routines can use.  The first routine is to
make a Diffie-Hellman private key pair:
\index{dh\_make\_key()}
\begin{verbatim}
int dh_make_key(prng_state *prng, int wprng, 
                int keysize, dh_key *key);
\end{verbatim}
The ``keysize'' is the size of the modulus you want in bytes.  Currently support sizes are 96 to 512 bytes which correspond 
to key sizes of 768 to 4096 bits. The smaller the key the faster it is to use however it will be less secure.  When 
specifying a size not explicitly supported by the library it will round {\em up} to the next key size.  If the size is 
above 512 it will return an error.  So if you pass ``keysize == 32'' it will use a 768 bit key but if you pass 
``keysize == 20000'' it will return an error.  The primes and generators used are built-into the library and were designed 
to meet very specific goals.  The primes are strong primes which means that if $p$ is the prime then
$p-1$ is equal to $2r$ where $r$ is a large prime.  The bases are chosen to generate a group of order $r$ to prevent
leaking a bit of the key.  This means the bases generate a very large prime order group which is good to make cryptanalysis
hard.

The next two routines are for exporting/importing Diffie-Hellman keys in a binary format.  This is useful for transport
over communication mediums.  

\index{dh\_export()} \index{dh\_import()}
\begin{verbatim}
int dh_export(unsigned char *out, unsigned long *outlen, 
              int type, dh_key *key);

int dh_import(const unsigned char *in, unsigned long inlen, dh_key *key);
\end{verbatim}

These two functions work just like the ``rsa\_export()'' and ``rsa\_import()'' functions except these work with 
Diffie-Hellman keys. Its important to note you do not have to free the ram for a ``dh\_key'' if an import fails.  You can free a 
``dh\_key'' using:
\begin{verbatim}
void dh_free(dh_key *key);
\end{verbatim}
After you have exported a copy of your public key (using {\bf PK\_PUBLIC} as ``type'') you can now create a shared secret 
with the other user using:
\index{dh\_shared\_secret()}
\begin{verbatim}
int dh_shared_secret(dh_key *private_key, 
                     dh_key *public_key, 
                     unsigned char *out, unsigned long *outlen);
\end{verbatim}

Where ``private\_key'' is the key you made and ``public\_key'' is the copy of the public key the other user sent you.  The result goes
into ``out'' and the length into ``outlen''.  If all went correctly the data in ``out'' should be identical for both parties.  It is important to
note that the two keys have to be the same size in order for this to work.  There is a function to get the size of a
key:
\index{dh\_get\_size()}
\begin{verbatim}
int dh_get_size(dh_key *key);
\end{verbatim}
This returns the size in bytes of the modulus chosen for that key.

\subsection{Remarks on Usage}
Its important that you hash the shared key before trying to use it as a key for a symmetric cipher or something.  An 
example program that communicates over sockets, using MD5 and 1024-bit DH keys is\footnote{This function is a small example.  It is suggested that proper packaging be used.  For example, if the public key sent is truncated these routines will not detect that.}:
\newpage
\begin{small}
\begin{verbatim}
int establish_secure_socket(int sock, int mode, unsigned char *key, 
                            prng_state *prng, int wprng)
{
   unsigned char buf[4096], buf2[4096];
   unsigned long x, len;
   int res, err, inlen;
   dh_key mykey, theirkey;

   /* make up our private key */
   if ((err = dh_make_key(prng, wprng, 128, &mykey)) != CRYPT_OK)  {
      return err;
   }

   /* export our key as public */ 
   x = sizeof(buf);
   if ((err = dh_export(buf, &x, PK_PUBLIC, &mykey)) != CRYPT_OK) {
      res = err;
      goto done2;
   }

   if (mode == 0) {
      /* mode 0 so we send first */
      if (send(sock, buf, x, 0) != x) {
         res = CRYPT_ERROR;
         goto done2;
      }          

      /* get their key */
      if ((inlen = recv(sock, buf2, sizeof(buf2), 0)) <= 0) {
         res = CRYPT_ERROR;
         goto done2;
      }
   } else {
      /* mode >0 so we send second */
      if ((inlen = recv(sock, buf2, sizeof(buf2), 0)) <= 0) {
         res = CRYPT_ERROR;
         goto done2;
      }

      if (send(sock, buf, x, 0) != x) {
         res = CRYPT_ERROR;
         goto done2;
      }
   }

   if ((err = dh_import(buf2, inlen, &theirkey)) != CRYPT_OK) { 
      res = err;
      goto done2;
   }

   /* make shared secret */
   x = sizeof(buf);
   if ((err = dh_shared_secret(&mykey, &theirkey, buf, &x)) != CRYPT_OK) {
      res = err;
      goto done;
   }
 
   /* hash it */
   len = 16;        /* default is MD5 so "key" must be at least 16 bytes long */
   if ((err = hash_memory(find_hash("md5"), buf, x, key, &len)) != CRYPT_OK) {
      res = err;
      goto done;
   }

   /* clean up and return */
   res = CRYPT_OK;
done:
   dh_free(&theirkey);
done2:
   dh_free(&mykey);
   zeromem(buf,  sizeof(buf));
   zeromem(buf2, sizeof(buf2));
   return res;
}
\end{verbatim}
\end{small}
\newpage
\subsection{Remarks on The Snippet}
When the above code snippet is done (assuming all went well) their will be a shared 128-bit key in the ``key'' array
passed to ``establish\_secure\_socket()''.

\section{Other Diffie-Hellman Functions}
In order to test the Diffie-Hellman function internal workings (e.g. the primes and bases) their is a test function made
available:
\index{dh\_test()}
\begin{verbatim}
int dh_test(void);
\end{verbatim}

This function returns {\bf CRYPT\_OK} if the bases and primes in the library are correct.  There is one last helper 
function:
\index{dh\_sizes()}
\begin{verbatim}
void dh_sizes(int *low, int *high);
\end{verbatim}
Which stores the smallest and largest key sizes support into the two variables.

\section{DH Packet}
Similar to the RSA related functions there are functions to encrypt or decrypt symmetric keys using the DH public key
algorithms.  
\index{dh\_encrypt\_key()} \index{dh\_decrypt\_key()}
\begin{verbatim}
int dh_encrypt_key(const unsigned char *inkey, unsigned long keylen,
                         unsigned char *out,  unsigned long *len, 
                         prng_state *prng, int wprng, int hash, 
                         dh_key *key);

int dh_decrypt_key(const unsigned char *in, unsigned long inlen,
                         unsigned char *outkey, unsigned long *keylen, 
                         dh_key *key);
\end{verbatim}
Where ``inkey'' is an input symmetric key of no more than 32 bytes.  Essentially these routines created a random public key
and find the hash of the shared secret.  The message digest is than XOR'ed against the symmetric key.  All of the 
required data is placed in ``out'' by ``dh\_encrypt\_key()''.   The hash must produce a message digest at least as large
as the symmetric key you are trying to share.

Similar to the RSA system you can sign and verify a hash of a message.
\index{dh\_sign\_hash()} \index{dh\_verify\_hash()}
\begin{verbatim}
int dh_sign_hash(const unsigned char *in,  unsigned long inlen,
                       unsigned char *out, unsigned long *outlen,
                       prng_state *prng, int wprng, dh_key *key);

int dh_verify_hash(const unsigned char *sig, unsigned long siglen,
                         const unsigned char *hash, unsigned long hashlen, 
                         int *stat, dh_key *key);
\end{verbatim}

The ``dh\_sign\_hash'' function signs the message hash in ``in'' of length ``inlen'' and forms a DH packet in ``out''.  
The ``dh\_verify\_hash'' function verifies the DH signature in ``sig'' against the hash in ``hash''.  It sets ``stat''
to non-zero if the signature passes or zero if it fails.

\chapter{Elliptic Curve Cryptography}

\section{Background}
The library provides a set of core ECC functions as well that are designed to be the Elliptic Curve analogy of all of the 
Diffie-Hellman routines in the previous chapter.  Elliptic curves (of certain forms) have the benefit that they are harder
to attack (no sub-exponential attacks exist unlike normal DH crypto) in fact the fastest attack requires the square root
of the order of the base point in time.  That means if you use a base point of order $2^{192}$ (which would represent a
192-bit key) then the work factor is $2^{96}$ in order to find the secret key.

The curves in this library are taken from the following website:
\begin{verbatim}
http://csrc.nist.gov/cryptval/dss.htm
\end{verbatim}

They are all curves over the integers modulo a prime.  The curves have the basic equation that is:
\begin{equation}
y^2 = x^3 - 3x + b\mbox{ }(\mbox{mod }p)
\end{equation}

The variable $b$ is chosen such that the number of points is nearly maximal.  In fact the order of the base points $\beta$ 
provided are very close to $p$ that is $\vert \vert \phi(\beta) \vert \vert \approx \vert \vert p \vert \vert$.  The curves
range in order from $\approx 2^{192}$ points to $\approx 2^{521}$.  According to the source document any key size greater
than or equal to 256-bits is sufficient for long term security.  

\section{Core Functions}

Like the DH routines there is a key structure ``ecc\_key'' used by the functions.  There is a function to make a key:
\index{ecc\_make\_key()}
\begin{verbatim}
int ecc_make_key(prng_state *prng, int wprng, 
                 int keysize, ecc_key *key);
\end{verbatim}

The ``keysize'' is the size of the modulus in bytes desired.  Currently directly supported values are 20, 24, 28, 32, 48 and 65 bytes which
correspond to key sizes of 160, 192, 224, 256, 384 and 521 bits respectively.  If you pass a key size that is between any key size
it will round the keysize up to the next available one.  The rest of the parameters work like they do in the ``dh\_make\_key()'' function.  
To free the ram allocated by a key call:
\index{ecc\_free()}
\begin{verbatim}
void ecc_free(ecc_key *key);
\end{verbatim}

To import and export a key there are: 
\index{ecc\_export()}
\index{ecc\_import()}
\begin{verbatim}
int ecc_export(unsigned char *out, unsigned long *outlen, 
               int type, ecc_key *key);

int ecc_import(const unsigned char *in, unsigned long inlen, ecc_key *key);
\end{verbatim}
These two work exactly like there DH counterparts.  Finally when you share your public key you can make a shared secret
with:
\index{ecc\_shared\_secret()}
\begin{verbatim}
int ecc_shared_secret(ecc_key *private_key, 
                      ecc_key *public_key, 
                      unsigned char *out, unsigned long *outlen);
\end{verbatim}
Which works exactly like the DH counterpart, the ``private\_key'' is your own key and ``public\_key'' is the key the other
user sent you.   Note that this function stores both $x$ and $y$ co-ordinates of the shared
elliptic point.  You should hash the output to get a shared key in a more compact and useful form (most of the entropy is 
in $x$ anyways).  Both keys have to be the same size for this to work, to help there is a function to get the size in bytes
 of a key.
\index{ecc\_get\_size()}
\begin{verbatim}
int ecc_get_size(ecc_key *key);
\end{verbatim}

To test the ECC routines and to get the minimum and maximum key sizes there are these two functions:
\index{ecc\_test()}
\begin{verbatim}
int ecc_test(void);
void ecc_sizes(int *low, int *high);
\end{verbatim}
Which both work like their DH counterparts.

\section{ECC Packet}
Similar to the RSA API there are two functions which encrypt and decrypt symmetric keys using the ECC public key
algorithms.

\index{ecc\_encrypt\_key()} \index{ecc\_decrypt\_key()}
\begin{verbatim}
int ecc_encrypt_key(const unsigned char *inkey, unsigned long keylen,
                          unsigned char *out,  unsigned long *len, 
                          prng_state *prng, int wprng, int hash, 
                          ecc_key *key);

int ecc_decrypt_key(const unsigned char *in, unsigned long inlen,
                          unsigned char *outkey, unsigned long *keylen, 
                          ecc_key *key);
\end{verbatim}

Where ``inkey'' is an input symmetric key of no more than 32 bytes.  Essentially these routines created a random public key
and find the hash of the shared secret.  The message digest is than XOR'ed against the symmetric key.  All of the required
data is placed in ``out'' by ``ecc\_encrypt\_key()''.   The hash chosen must produce a message digest at least as large
as the symmetric key you are trying to share.

There are also functions to sign and verify the hash of a message.
\index{ecc\_sign\_hash()} \index{ecc\_verify\_hash()}
\begin{verbatim}
int ecc_sign_hash(const unsigned char *in,  unsigned long inlen,
                        unsigned char *out, unsigned long *outlen,
                        prng_state *prng, int wprng, ecc_key *key);

int ecc_verify_hash(const unsigned char *sig, unsigned long siglen,
                    const unsigned char *hash, unsigned long hashlen, 
                          int *stat, ecc_key *key);
\end{verbatim}

The ``ecc\_sign\_hash'' function signs the message hash in ``in'' of length ``inlen'' and forms a ECC packet in ``out''.  
The ``ecc\_verify\_hash'' function verifies the ECC signature in ``sig'' against the hash in ``hash''.  It sets ``stat''
to non-zero if the signature passes or zero if it fails.


\section{ECC Keysizes}
With ECC if you try and sign a hash that is bigger than your ECC key you can run into problems.  The math will still work
and in effect the signature will still work.  With ECC keys the strength of the signature is limited by the size of
the hash or the size of they key, whichever is smaller.  For example, if you sign with SHA256 and a ECC-160 key in effect
you have 160-bits of security (e.g. as if you signed with SHA-1).  

The library will not warn you if you make this mistake so it is important to check yourself before using the 
signatures.

\chapter{Digital Signature Algorithm}
\section{Introduction}
The Digital Signature Algorithm (or DSA) is a variant of the ElGamal Signature scheme which has been modified to 
reduce the bandwidth of a signature.  For example, to have ``80-bits of security'' with ElGamal you need a group of 
order at least 1024-bits.  With DSA you need a group of order at least 160-bits.  By comparison the ElGamal signature
would require at least 256 bytes where as the DSA signature would require only at least 40 bytes.  

The API for the DSA is essentially the same as the other PK algorithms.  Except in the case of DSA no encryption or
decryption routines are provided.  

\section{Key Generation}
To make a DSA key you must call the following function
\begin{verbatim}
int dsa_make_key(prng_state *prng, int wprng, 
                 int group_size, int modulus_size, 
                 dsa_key *key);
\end{verbatim}
The variable ``prng'' is an active PRNG state and ``wprng'' the index to the descriptor.  ``group\_size'' and 
``modulus\_size'' control the difficulty of forging a signature.  Both parameters are in bytes.  The larger the
``group\_size'' the more difficult a forgery becomes upto a limit.  The value of $group\_size$ is limited by 
$15 < group\_size < 1024$ and $modulus\_size - group\_size < 512$.  Suggested values for the pairs are as follows.

\begin{center}
\begin{tabular}{|c|c|c|}
\hline \textbf{Bits of Security} & \textbf{group\_size} & \textbf{modulus\_size} \\
\hline 80  & 20 & 128 \\
\hline 120 & 30 & 256 \\
\hline 140 & 35 & 384 \\
\hline 160 & 40 & 512 \\
\hline
\end{tabular}
\end{center}

When you are finished with a DSA key you can call the following function to free the memory used.
\index{dsa\_free()}
\begin{verbatim}
void dsa_free(dsa_key *key);
\end{verbatim}

\section{Key Verification}
Each DSA key is composed of the following variables.

\begin{enumerate}
  \item $q$ a small prime of magnitude $256^{group\_size}$.  
  \item $p = qr + 1$ a large prime of magnitude $256^{modulus\_size}$ where $r$ is a random even integer.
  \item $g = h^r \mbox{ (mod }p\mbox{)}$ a generator of order $q$ modulo $p$.  $h$ can be any non-trivial random 
        value.  For this library they start at $h = 2$ and step until $g$ is not $1$.
  \item $x$ a random secret (the secret key) in the range $1 < x < q$ 
  \item $y = g^x \mbox{ (mod }p\mbox{)}$ the public key.
\end{enumerate}

A DSA key is considered valid if it passes all of the following tests.

\begin{enumerate}
   \item $q$ must be prime.
   \item $p$ must be prime.
   \item $g$ cannot be one of $\lbrace -1, 0, 1 \rbrace$ (modulo $p$).
   \item $g$ must be less than $p$.
   \item $(p-1) \equiv 0 \mbox{ (mod }q\mbox{)}$.
   \item $g^q \equiv 1 \mbox{ (mod }p\mbox{)}$.
   \item $1 < y < p - 1$
   \item $y^q \equiv 1 \mbox{ (mod }p\mbox{)}$.
\end{enumerate}

Tests one and two ensure that the values will at least form a field which is required for the signatures to  
function.  Tests three and four ensure that the generator $g$ is not set to a trivial value which would make signature
forgery easier.  Test five ensures that $q$ divides the order of multiplicative sub-group of $\Z/p\Z$. Test six
ensures that the generator actually generates a prime order group.  Tests seven and eight ensure that the public key
is within range and belongs to a group of prime order.  Note that test eight does not prove that $g$ generated $y$ only
that $y$ belongs to a multiplicative sub-group of order $q$. 

The following function will perform these tests.

\index{dsa\_verify\_key()}
\begin{verbatim}
int dsa_verify_key(dsa_key *key, int *stat);
\end{verbatim}

This will test ``key'' and store the result in ``stat''.  If the result is $stat = 0$ the DSA key failed one of the tests
and should not be used at all.  If the result is $stat = 1$ the DSA key is valid (as far as valid mathematics are concerned).

\section{Signatures}
To generate a DSA signature call the following function

\index{dsa\_sign\_hash()}
\begin{verbatim}
int dsa_sign_hash(const unsigned char *in,  unsigned long inlen,
                        unsigned char *out, unsigned long *outlen,
                        prng_state *prng, int wprng, dsa_key *key);
\end{verbatim}

Which will sign the data in ``in'' of length ``inlen'' bytes.  The signature is stored in ``out'' and the size
of the signature in ``outlen''.  If the signature is longer than the size you initially specify in ``outlen'' nothing
is stored and the function returns an error code.  The DSA ``key'' must be of the \textbf{PK\_PRIVATE} persuasion.

To verify a hash created with that function use the following function

\index{dsa\_verify\_hash()} 
\begin{verbatim}
int dsa_verify_hash(const unsigned char *sig, unsigned long siglen,
                    const unsigned char *hash, unsigned long inlen, 
                    int *stat, dsa_key *key);
\end{verbatim}
Which will verify the data in ``hash'' of length ``inlen'' against the signature stored in ``sig'' of length ``siglen''.  
It will set ``stat'' to $1$ if the signature is valid, otherwise it sets ``stat'' to $0$.  

\section{Import and Export}

To export a DSA key so that it can be transported use the following function
\index{dsa\_export()}
\begin{verbatim}
int dsa_export(unsigned char *out, unsigned long *outlen, 
               int type, 
               dsa_key *key);
\end{verbatim}
This will export the DSA ``key'' to the buffer ``out'' and set the length in ``outlen'' (which must have been previously
initialized to the maximum buffer size).  The ``type`` variable may be either \textbf{PK\_PRIVATE} or \textbf{PK\_PUBLIC}
depending on whether you want to export a private or public copy of the DSA key.

To import an exported DSA key use the following function

\index{dsa\_import()}
\begin{verbatim}
int dsa_import(const unsigned char *in, unsigned long inlen, 
               dsa_key *key);
\end{verbatim}

This will import the DSA key from the buffer ``in'' of length ``inlen'' to the ``key''.  If the process fails the function
will automatically free all of the heap allocated in the process (you don't have to call dsa\_free()).  

\chapter{Standards Support}
\section{DER Support}
DER or ``Distinguished Encoding Rules'' is a subset of the ASN.1 encoding rules that is fully deterministic and
ideal for cryptography.  In particular ASN.1 specifies an INTEGER type for storing arbitrary sized integers.  DER
further limits the ASN.1 specifications to a deterministic encoding.

\subsection{Storing INTEGER types}
\index{der\_encode\_integer()}
\begin{alltt}
int der_encode_integer(mp_int *num, unsigned char *out, unsigned long *outlen);
\end{alltt}

This will store the integer in ``num'' to the output buffer ``out'' of length ``outlen''.  It only stores
non--negative numbers.  It stores the number of octets used back in ``outlen''.

\subsection{Reading INTEGER types}
\index{der\_decode\_integer()}
\begin{alltt}
int der_decode_integer(const unsigned char *in, unsigned long *inlen, mp_int *num);
\end{alltt}
This will decode the DER encoded INTEGER in ``in'' of length ``inlen'' and store the resulting integer
in ``num''.  It will store the bytes read in ``inlen'' which is handy if you have to parse multiple
data items out of a binary packet.

\subsection{INTEGER length}
\index{der\_length\_integer()}
\begin{alltt}
int der_length_integer(mp_int *num, unsigned long *len);
\end{alltt}
This will determine the length of the DER encoding of the integer ``num'' and store it in ``len''.

\subsection{Multiple INTEGER types}
To simplify the DER encoding/decoding there are two functions two handle multple types at once.

\index{der\_put\_multi\_integer()}
\index{der\_get\_multi\_integer()}
\begin{alltt}
int der_put_multi_integer(unsigned char *dst, unsigned long *outlen, mp_int *num, ...);
int der_get_multi_integer(const unsigned char *src, unsigned long *inlen,  mp_int *num, ...);
\end{alltt}

These will handle multiple encodings/decodings at once.  They work like their single operand counterparts
except they handle a \textbf{NULL} terminated list of operands.

\begin{verbatim}
#include <mycrypt.h>
int main(void)
{
   mp_int        a, b, c, d;
   unsigned char buffer[1000];
   unsigned long len;
   int           err;

   /* init a,b,c,d with some values ... */

   /* ok we want to store them now... */
   len = sizeof(buffer);
   if ((err = der_put_multi_integer(buffer, &len, 
                                    &a, &b, &c, &d, NULL)) != CRYPT_OK) {
      // error
   }
   printf("I stored %lu bytes in buf\n", len);

   /* ok say we want to get them back for fun */
   /* len set previously...otherwise set it to the size of the packet */
   if ((err = der_get_multi_integer(buffer, &len,
                                    &a, &b, &c, &d, NULL)) != CRYPT_OK) {
      // error
   }
   printf("I read %lu bytes from buf\n", len);
}
\end{verbatim}
\section{Password Based Cryptography}
\subsection{PKCS \#5}
In order to securely handle user passwords for the purposes of creating session keys and chaining IVs the PKCS \#5 was drafted.   PKCS \#5
is made up of two algorithms, Algorithm One and Algorithm Two.  Algorithm One is the older fairly limited algorithm which has been implemented
for completeness.  Algorithm Two is a bit more modern and more flexible to work with.

\subsection{Algorithm One}
Algorithm One accepts as input a password, an 8--byte salt and an iteration counter.  The iteration counter is meant to act as delay for
people trying to brute force guess the password.  The higher the iteration counter the longer the delay.  This algorithm also requires a hash 
algorithm and produces an output no longer than the output of the hash.  

\index{pkcs\_5\_alg1()}
\begin{alltt}
int pkcs_5_alg1(const unsigned char *password, unsigned long password_len, 
                const unsigned char *salt, 
                int iteration_count,  int hash_idx,
                unsigned char *out,   unsigned long *outlen)
\end{alltt}
Where ``password'' is the users password.  Since the algorithm allows binary passwords you must also specify the length in ``password\_len''.  
The ``salt'' is a fixed size 8--byte array which should be random for each user and session.  The ``iteration\_count'' is the delay desired
on the password.  The ``hash\_idx'' is the index of the hash you wish to use in the descriptor table.  

The output of length upto ``outlen'' is stored in ``out''.  If ``outlen'' is initially larger than the size of the hash functions output
it is set to the number of bytes stored.  If it is smaller than not all of the hash output is stored in ``out''.

\subsection{Algorithm Two}

Algorithm Two is the recommended algorithm for this task.  It allows variable length salts and can produce outputs larger than the 
hash functions output.  As such it can easily be used to derive session keys for ciphers and MACs as well initial vectors as required
from a single password and invokation of this algorithm.

\index{pkcs\_5\_alg2()}
\begin{alltt}
int pkcs_5_alg2(const unsigned char *password, unsigned long password_len, 
                const unsigned char *salt,     unsigned long salt_len,
                int iteration_count,           int hash_idx,
                unsigned char *out,            unsigned long *outlen)
\end{alltt}
Where ``password'' is the users password.  Since the algorithm allows binary passwords you must also specify the length in ``password\_len''.  
The ``salt'' is an array of size ``salt\_len''.  It should be random for each user and session.  The ``iteration\_count'' is the delay desired
on the password.  The ``hash\_idx'' is the index of the hash you wish to use in the descriptor table.   The output of length upto 
``outlen'' is stored in ``out''.

\begin{alltt}
/* demo to show how to make session state material from a password */
#include <mycrypt.h>
int main(void)
\{
    unsigned char password[100], salt[100],
                  cipher_key[16], cipher_iv[16],
                  mac_key[16], outbuf[48];
    int           err, hash_idx;
    unsigned long outlen, password_len, salt_len;

    /* register hash and get it's idx .... */

    /* get users password and make up a salt ... */

    /* create the material (100 iterations in algorithm) */
    outlen = sizeof(outbuf);
    if ((err = pkcs_5_alg2(password, password_len, salt, salt_len, 
                           100, hash_idx, outbuf, &outlen)) != CRYPT_OK) \{
       /* error handle */
    \}

    /* now extract it */
    memcpy(cipher_key, outbuf, 16);
    memcpy(cipher_iv,  outbuf+16, 16);
    memcpy(mac_key,    outbuf+32, 16);

    /* use material (recall to store the salt in the output) */
\}
\end{alltt}


\chapter{Miscellaneous}
\section{Base64 Encoding and Decoding}
The library provides functions to encode and decode a RFC1521 base64 coding scheme.  This means that it can decode what it 
encodes but the format used does not comply to any known standard.  The characters used in the mappings are:
\begin{verbatim}
ABCDEFGHIJKLMNOPQRSTUVWXYZabcdefghijklmnopqrstuvwxyz0123456789+/
\end{verbatim}
Those characters should are supported in virtually any 7-bit ASCII system which means they can be used for transport over
common e-mail, usenet and HTTP mediums.  The format of an encoded stream is just a literal sequence of ASCII characters
where a group of four represent 24-bits of input.  The first four chars of the encoders output is the length of the 
original input.  After the first four characters is the rest of the message.

Often it is desirable to line wrap the output to fit nicely in an e-mail or usenet posting.  The decoder allows you to
put any character (that is not in the above sequence) in between any character of the encoders output.  You may not however,
break up the first four characters.

To encode a binary string in base64 call:
\index{base64\_encode()}  \index{base64\_decode()} 
\begin{verbatim}
int base64_encode(const unsigned char *in, unsigned long len, 
                  unsigned char *out, unsigned long *outlen);
\end{verbatim}
Where ``in'' is the binary string and ``out'' is where the ASCII output is placed.  You must set the value of ``outlen'' prior
to calling this function and it sets the length of the base64 output in ``outlen'' when it is done.  To decode a base64 
string call:
\begin{verbatim}
int base64_decode(const unsigned char *in, unsigned long len, 
                  unsigned char *out, unsigned long *outlen);
\end{verbatim}

\section{The Multiple Precision Integer Library (MPI)}
The library comes with a copy of LibTomMath  which is a multiple precision integer library written by the
author of LibTomCrypt.  LibTomMath is a trivial to use ANSI C compatible large integer library which is free 
for all uses and is distributed freely.

At the heart of all the functions is the data type ``mp\_int'' (defined in tommath.h).  This data type is what 
will hold all large integers.  In order to use an mp\_int one must initialize it first, for example:
\begin{verbatim}
#include <mycrypt.h> /* mycrypt.h includes mpi.h automatically */
int main(void)
{ 
   mp_int bignum;
   
   /* initialize it */
   mp_init(&bignum);

   return 0;
}
\end{verbatim}
If you are unfamiliar with the syntax of C the \& symbol is used to pass the address of ``bignum'' to the function.  All
LibTomMath functions require the address of the parameters.  To free the memory of a mp\_int use (for example):
\begin{verbatim}
mp_clear(&bignum);
\end{verbatim}

The functions also have the basic form of one of the following:
\begin{verbatim}
mp_XXX(mp_int *a);
mp_XXX(mp_int *a, mp_int *b, mp_int *c);
mp_XXX(mp_int *a, mp_int *b, mp_int *c, mp_int *d);
\end{verbatim}

Where they perform some operation and store the result in the mp\_int variable passed on the far right.  
For example, to compute $c = a + b \mbox{ }(\mbox{mod }m)$ you would call:
\begin{verbatim}
mp_addmod(&a, &b, &m, &c);
\end{verbatim}

\subsection{Binary Forms of ``mp\_int'' Variables}

Often it is required to store a ``mp\_int'' in binary form for transport (e.g. exporting a key, packet 
encryption, etc.).  LibTomMath includes two functions to help when exporting numbers:
\begin{verbatim}
int mp_raw_size(mp_int *num);
mp_toraw(&num, buf);
\end{verbatim}

The former function gives the size in bytes of the raw format and the latter function actually stores the raw data.  All
``mp\_int'' numbers are stored in big endian form (like PKCS demands) with the first byte being the sign of the number.  The
``rsa\_exptmod()'' function differs slightly since it will take the input in the form exactly as PKCS demands (without the
leading sign byte).  All other functions include the sign byte (since its much simpler just to include it).  The sign byte
must be zero for positive numbers and non-zero for negative numbers.  For example,
the sequence:
\begin{verbatim}
00 FF 30 04
\end{verbatim}
Represents the integer $255 \cdot 256^2 + 48 \cdot 256^1 + 4 \cdot 256^0$ or 16,723,972.

To read a binary string back into a ``mp\_int'' call:
\begin{verbatim}
mp_read_raw(mp_int *num, unsigned char *str, int len);
\end{verbatim}
Where ``num'' is where to store it, ``str'' is the binary string (including the leading sign byte) and ``len'' is the 
length of the binary string.

\subsection{Primality Testing}
\index{Primality Testing}
The library includes primality testing and random prime functions as well.  The primality tester will perform the test in
two phases.  First it will perform trial division by the first few primes.  Second it will perform eight rounds of the 
Rabin-Miller primality testing algorithm.  If the candidate passes both phases it is declared prime otherwise it is declared
composite.  No prime number will fail the two phases but composites can.  Each round of the Rabin-Miller algorithm reduces
the probability of a pseudo-prime by $1 \over 4$ therefore after sixteen rounds the probability is no more than 
$\left ( { 1 \over 4 } \right )^{8} = 2^{-16}$.  In practice the probability of error is in fact much lower than that.

When making random primes the trial division step is in fact an optimized implementation of ``Implementation of Fast RSA Key Generation on Smart Cards''\footnote{Chenghuai Lu, Andre L. M. dos Santos and Francisco R. Pimentel}.
In essence a table of machine-word sized residues are kept of a candidate modulo a set of primes.  When the candiate
is rejected and ultimately incremented to test the next number the residues are updated without using multi-word precision
math operations.  As a result the routine can scan ahead to the next number required for testing with very little work
involved.

In the event that a composite did make it through it would most likely cause the the algorithm trying to use it to fail.  For 
instance, in RSA two primes $p$ and $q$ are required.  The order of the multiplicative sub-group (modulo $pq$) is given 
as $\phi(pq)$ or $(p - 1)(q - 1)$.  The decryption exponent $d$ is found as $de \equiv 1\mbox{ }(\mbox{mod } \phi(pq))$.  If either $p$ or $q$ is composite the value of $d$ will be incorrect and the user
will not be able to sign or decrypt messages at all.  Suppose $p$ was prime and $q$ was composite this is just a variation of 
the multi-prime RSA.  Suppose $q = rs$ for two primes $r$ and $s$ then $\phi(pq) = (p - 1)(r - 1)(s - 1)$ which clearly is 
not equal to $(p - 1)(rs - 1)$.

These are not technically part of the LibTomMath library but this is the best place to document them.  
To test if a ``mp\_int'' is prime call:
\begin{verbatim}
int is_prime(mp_int *N, int *result);
\end{verbatim}
This puts a one in ``result'' if the number is probably prime, otherwise it places a zero in it.  It is assumed that if 
it returns an error that the value in ``result'' is undefined.  To make 
a random prime call:
\begin{verbatim}
int rand_prime(mp_int *N, unsigned long len, prng_state *prng, int wprng);
\end{verbatim}
Where ``len'' is the size of the prime in bytes ($2 \le len \le 256$).  You can set ``len'' to the negative size you want
to get a prime of the form $p \equiv 3\mbox{ }(\mbox{mod } 4)$.  So if you want a 1024-bit prime of this sort pass 
``len = -128'' to the function.  Upon success it will return {\bf CRYPT\_OK} and ``N'' will contain an integer which
is very likely prime.

\chapter{Programming Guidelines}

\section{Secure Pseudo Random Number Generators}
Probably the singal most vulnerable point of any cryptosystem is the PRNG.  Without one generating and protecting secrets
would be impossible.  The requirement that one be setup correctly is vitally important and to address this point the library
does provide two RNG sources that will address the largest amount of end users as possible.  The ``sprng'' PRNG provided 
provides and easy to access source of entropy for any application on a *NIX or Windows computer.  

However, when the end user is not on one of these platforms the application developer must address the issue of finding
entropy.  This manual is not designed to be a text on cryptography.  I would just like to highlight that when you design
a cryptosystem make sure the first problem you solve is getting a fresh source of entropy.  

\section{Preventing Trivial Errors}
Two simple ways to prevent trivial errors is to prevent overflows and to check the return values.  All of the functions
which output variable length strings will require you to pass the length of the destination.  If the size of your output
buffer is smaller than the output it will report an error.  Therefore, make sure the size you pass is correct!

Also virtually all of the functions return an error code or {\bf CRYPT\_OK}.  You should detect all errors as simple 
typos or such can cause algorithms to fail to work as desired.

\section{Registering Your Algorithms}
To avoid linking and other runtime errors it is important to register the ciphers, hashes and PRNGs you intend to use 
before you try to use them.  This includes any function which would use an algorithm indirectly through a descriptor table.

A neat bonus to the registry system is that you can add external algorithms that are not part of the library without 
having to hack the library.  For example, suppose you have a hardware specific PRNG on your system.  You could easily 
write the few functions required plus a descriptor.  After registering your PRNG all of the library functions that 
need a PRNG can instantly take advantage of it.

\section{Key Sizes}

\subsection{Symmetric Ciphers}
For symmetric ciphers use as large as of a key as possible.  For the most part ``bits are cheap'' so using a 256-bit key
is not a hard thing todo.  

\subsection{Assymetric Ciphers}
The following chart gives the work factor for solving a DH/RSA public key using the NFS.  The work factor for a key of order
$n$ is estimated to be
\begin{equation}
e^{1.923 \cdot ln(n)^{1 \over 3} \cdot ln(ln(n))^{2 \over 3}} 
\end{equation}

Note that $n$ is not the bit-length but the magnitude.  For example, for a 1024-bit key $n = 2^{1024}$.  The work required 
is:
\begin{center}
\begin{tabular}{|c|c|}
    \hline RSA/DH Key Size (bits) & Work Factor ($log_2$) \\
    \hline 512 & 63.92 \\
    \hline 768 & 76.50 \\
    \hline 1024 & 86.76 \\
    \hline 1536 & 103.37 \\
    \hline 2048 & 116.88 \\
    \hline 2560 & 128.47 \\
    \hline 3072 & 138.73 \\
    \hline 4096 & 156.49 \\
    \hline 
\end{tabular}
\end{center}

The work factor for ECC keys is much higher since the best attack is still fully exponentional.  Given a key of magnitude
$n$ it requires $\sqrt n$ work.  The following table sumarizes the work required:
\begin{center}
\begin{tabular}{|c|c|}
    \hline ECC Key Size (bits) & Work Factor ($log_2$) \\
    \hline 160 & 80  \\
    \hline 192 & 96  \\
    \hline 224 & 112 \\
    \hline 256 & 128 \\
    \hline 384 & 192 \\
    \hline 521 & 260.5 \\
    \hline
\end{tabular}
\end{center}

Using the above tables the following suggestions for key sizes seems appropriate:
\begin{center}
\begin{tabular}{|c|c|c|}
    \hline Security Goal & RSA/DH Key Size (bits) & ECC Key Size (bits) \\
    \hline Short term (less than a year) & 1024 & 160 \\
    \hline Short term (less than five years) & 1536 & 192 \\
    \hline Long Term (less than ten years) & 2560 & 256 \\
    \hline 
\end{tabular}
\end{center}

\section{Thread Safety}
The library is not thread safe but several simple precautions can be taken to avoid any problems.  The registry functions
such as register\_cipher() are not thread safe no matter what you do.  Its best to call them from your programs initializtion
code before threads are initiated.

The rest of the code uses state variables you must pass it such as hash\_state, hmac\_state, etc.  This means that if each
thread has its own state variables then they will not affect each other.  This is fairly simple with symmetric ciphers
and hashes.  However, the keyring and PRNG support is something the threads will want to share.  The simplest workaround 
is create semaphores or mutexes around calls to those functions.  

Since C does not have standard semaphores this support is not native to Libtomcrypt.  Even a C based semaphore is not entire
possible as some compilers may ignore the ``volatile'' keyword or have multiple processors.  Provide your host application
is modular enough putting the locks in the right place should not bloat the code significantly and will solve all thread
safety issues within the library.

\chapter{Configuring and Building the Library}
\section{Introduction}
The library is fairly flexible about how it can be built, used and generally distributed.  Additions are being made with
each new release that will make the library even more flexible.  Each of the classes of functions can be disabled during
the build process to make a smaller library.  This is particularly useful for shared libraries.

\section{Building a Static Library}
The library can be built as a static library which is generally the simplest and most portable method of 
building the library.  With a CC or GCC equipped platform you can issue the following

\begin{alltt}
make install_lib
\end{alltt}

Which will build the library and install it in /usr/lib (as well as the headers in /usr/include).  The destination
directory of the library and headers can be changed by editing ``makefile''.  The variable LIBNAME controls
where the library is to be installed and INCNAME controls where the headers are to be installed.  A developer can 
then use the library by including ``mycrypt.h'' in their program and linking against ``libtomcrypt.a''.

A static library can also be built with the Intel C Compiler  (ICC) by issuing the following

\begin{alltt}
make -f makefile.icc install
\end{alltt}

This will also build ``libtomcrypt.a'' except that it will use ICC.  Additionally Microsoft's Visual C 6.00 can be used
by issuing

\begin{alltt}
nmake -f makefile.msvc
\end{alltt}

You will have to manually copy ``tomcrypt.lib'' and the headers to your MSVC lib/inc directories.

\subsection{MPI Control}
If you already have LibTomMath installed you can safely remove it from the build.  By commenting the line
in the appropriate makefile which starts with 

\begin{alltt}
MPIOBJECT=mpi
\end{alltt}

Simply place a \# at the start and re-build the library.  To properly link applications you will have to also
link in LibTomMath.  Removing MPI has the benefit of cutting down the library size as well potentially have access
to the latest mpi.

\section{Building a Shared Library}
LibTomCrypt can also be built as a shared library (.so, .dll, etc...).  With non-Windows platforms the assumption
of the presence of gcc and ``libtool'' has been made.  These are fairly common on Unix/Linux/BSD platforms.  To
build a .so shared library issue 

\begin{alltt}
make -f makefile.shared
\end{alltt}
This will use libtool and gcc to build a shared library ``libtomcrypt.la'' as well as a static library ``libtomcrypt.a''
and install them into /usr/lib (and the headers into /usr/include).  To link your application you should use the 
libtool program in ``--mode=link''.

You can also build LibTomCrypt as a shared library (DLL) in Windows with Cygwin.  Issue the following

\begin{alltt}
make -f makefile.cygwin_dll
\end{alltt}
This will build ``libtomcrypt.dll.a'' which is an import library for ``libtomcrypt.dll''.  You must copy 
``libtomcrypt.dll.a'' to your library directory, ``libtomcrypt.dll' to somewhere in your PATH and the header
files to your include directory.  So long as ``libtomcrypt.dll'' is in your system path you can run any LibTomCrypt
program that uses it.

\section{mycrypt\_cfg.h}
The file ``mycrypt\_cfg.h'' is what lets you control various high level macros which control the behaviour 
of the library. 

\subsubsection{ARGTYPE}
This lets you control how the \_ARGCHK macro will behave.  The macro is used to check pointers inside the functions against
NULL.  There are three settings for ARGTYPE.  When set to 0 it will have the default behaviour of printing a message to 
stderr and raising a SIGABRT signal.  This is provided so all platforms that use libtomcrypt can have an error that functions
similarly.  When set to 1 it will simply pass on to the assert() macro.  When set to 2 it will resolve to a empty macro
and no error checking will be performed.

\subsubsection{Endianess}
There are five macros related to endianess issues.  For little endian platforms define, ENDIAN\_LITTLE.  For big endian
platforms define ENDIAN\_BIG.  Similarly when the default word size of an ``unsigned long'' is 32-bits define ENDIAN\_32BITWORD
or define ENDIAN\_64BITWORD when its 64-bits.  If you do not define any of them the library will automatically use ENDIAN\_NEUTRAL
which will work on all platforms.

Currently LibTomCrypt will detect x86-32 and x86-64 running GCC as well as x86-32 running MSVC.  

\section{The Configure Script}
There are also options you can specify from the configure script or ``mycrypt\_custom.h''.  

\subsubsection{X memory routines}
At the top of mycrypt\_custom.h are four macros denoted as XMALLOC, XCALLOC, XREALLOC and XFREE which resolve to 
the name of the respective functions.  This lets you substitute in your own memory routines.  If you substitute in 
your own functions they must behave like the standard C library functions in terms of what they expect as input and 
output.  By default the library uses the standard C routines.

\subsubsection{X clock routines}
The rng\_get\_bytes() function can call a function that requires the clock() function.  These macros let you override
the default clock() used with a replacement.  By default the standard C library clock() function is used.

\subsubsection{NO\_FILE}
During the build if NO\_FILE is defined then any function in the library that uses file I/O will not call the file I/O 
functions and instead simply return CRYPT\_NOP.  This should help resolve any linker errors stemming from a lack of
file I/O on embedded platforms.

\subsubsection{CLEAN\_STACK}
When this functions is defined the functions that store key material on the stack will clean up afterwards.  
Assumes that you have no memory paging with the stack.

\subsubsection{LTC\_TEST}
When this has been defined the various self--test functions (for ciphers, hashes, prngs, etc) are included in the build.
When this has been undefined the tests are removed and if called will return CRYPT\_NOP.

\subsubsection{Symmetric Ciphers, One-way Hashes, PRNGS and Public Key Functions}
There are a plethora of macros for the ciphers, hashes, PRNGs and public key functions which are fairly 
self-explanatory.  When they are defined the functionality is included otherwise it is not.  There are some 
dependency issues which are noted in the file.  For instance, Yarrow requires CTR chaining mode, a block 
cipher and a hash function.

\subsubsection{TWOFISH\_SMALL and TWOFISH\_TABLES}
Twofish is a 128-bit symmetric block cipher that is provided within the library.  The cipher itself is flexible enough
to allow some tradeoffs in the implementation.  When TWOFISH\_SMALL is defined the scheduled symmetric key for Twofish 
requires only 200 bytes of memory.  This is achieved by not pre-computing the substitution boxes.  Having this 
defined will also greatly slow down the cipher.  When this macro is not defined Twofish will pre-compute the 
tables at a cost of 4KB of memory.  The cipher will be much faster as a result.  

When TWOFISH\_TABLES is defined the cipher will use pre-computed (and fixed in code) tables required to work.  This is
useful when TWOFISH\_SMALL is defined as the table values are computed on the fly.  When this is defined the code size
will increase by approximately 500 bytes.  If this is defined but TWOFISH\_SMALL is not the cipher will still work but
it will not speed up the encryption or decryption functions.

\subsubsection{SMALL\_CODE}
When this is defined some of the code such as the Rijndael and SAFER+ ciphers are replaced with smaller code variants.
These variants are slower but can save quite a bit of code space.

\section{MPI Tweaks}
\subsection{RSA Only Tweak}
If you plan on only using RSA with moduli in the range of 1024 to 2560 bits you can enable a series of tweaks
to reduce the library size.  Follow these steps

\begin{enumerate}
   \item Undefine MDSA, MECC and MDH from mycrypt\_custom.h
   \item Undefine LTM\_ALL  from tommath\_superclass.h
   \item Define SC\_RSA\_1 from tommath\_superclass.h
   \item Rebuild the library.
\end{enumerate}



\input{crypt.ind}

\end{document}
